### 2018-07-13 paper.tex
(*) First commit

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

\cite{ceu.tecs17}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{my}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2018-07-16 paper.tex
...

@@
 
+\usepackage{xspace}
+\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
+
 \begin{document}
@@
 %% Title information
-\title[Short Title]{Full Title}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-\subtitle{Subtitle}                     %% \subtitle is optional
-\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
+\title{Where Do Events Come From?}
+\subtitle{Reactive Programming From The Ground Up}
 
@@
 %% before \maketitle command
+
 \begin{abstract}
-Text of abstract \ldots.
+In reactive and event-based systems, execution is guided by an external
+environment which produces inputs and consumes output events to and from the
+applications.
+Reactive languages provide dedicated syntax and semantics to deal with events
+and greatly simplify the programming experience in this domain.
+%
+Nevertheless, the environment is typically prefabricated in a host language and
+the very central concept of events is implemented externally to the reactive
+language.
+This separation exposes the environment as a rigid system component that
+evolves in separate from the application.
+%
+In this work, we propose an interrupt handler primitive for a reactive language
+targeting embedded systems in order to take control of the whole event loop,
+from the hardware input source and back to the hardware output sink.
+%
+We propose the new primitive in the context of the structured reactive language
+\CEU and discuss how it synergizes with the structured paradigm to provide
+automatic standby for applications and hot swapping for drivers during runtime.
 \end{abstract}
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{interrupt service routine,
+          sleep mode,
+          synchronous reactive programming}
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
@@
 
-Text of paper \ldots
+Reactive applications interact continuously and in real time with the external
+world through input and output devices (e.g., buttons, displays, timers, etc.).
+%
+These interactions are typically represented as input events flowing from the
+devices to the application and as output events flowing from the application to
+the devices.
+%
+As illustrated in Figure~\ref{fig.loop}, devices can be encapsulated as a
+single component, \emph{the environment}, which controls the system execution
+in an event loop:
+the application sits idle waiting for an input;
+the environment awakes the application on the occurrence of an input;
+the application reacts to the input and generates back one or more outputs;
+the application becomes idle and the loop restarts.
 
-\cite{ceu.tecs17}
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{loop}
+\caption{ Event loop in reactive systems.
+          The environment controls the application through input \& output
+          events.
+\label{fig.loop}
+}
+\end{figure}
+
+The environment is typically implemented in a host language (e.g., C) and
+controls the main event loop, invoking entry points in the reactive language
+runtime on the occurrence of inputs and also receiving output calls, both
+through a documented API.
+%
+As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
+between the environment and the running program~\cite{esterel.book.compiling},
+while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
+out values to JavaScript as commands and listening for values as
+subscriptions~\cite{frp.elm.ports}.
+
+The event-based separation between the application and the environment is
+arguably inevitable and happens at the right level of abstraction, since it
+connects the application with the operating system resources through a
+non-invasive API.
+%
+However, it involves two languages and the programmer may have to unnecessarily
+deal with multiple syntaxes, incompatible type systems, and different address
+spaces.
+%
+Furthermore, in the context of embedded systems, a proper host OS may even be
+absent or lacking enough drivers, which requires more low-level intervention
+from the application.
+
+In this work, we propose
+an interrupt handler primitive for a reactive language in
+order to take control of the whole event loop, from the
+hardware input source and back to the hardware output
+sink. We propose the new primitive in the context of
+the structured reactive language \CEU and discuss how
+it synergizes with the structured paradigm to provide
+automatic standby for applications and hot swapping
+for drivers during runtime.
+
+with different syntaxes and semantics.
+For instance, , type systems, and
+
+abstraction
+    - rigid
+    - two languages
+        - syntax and semantics
+            - integrates with features such as lexical scope, type system
+    - platforms w/o OS
+
+\newpage
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{sync-async}
+\caption{ Event loop in reactive systems.
+          The environment controls the application through input \& output
+          events.
+\label{fig.loop}
+}
+\end{figure}
+
+what would be
+
+output (int,int) O;
+
+becomes
+
+output (int x, int y) O do
+end
+
+\begin{comment}
+% Esterel
+https://books.google.com.br/books?id=O5zi14i0KP4C&pg=PA31&lpg=PA31&dq=esterel+external+signal+api&source=bl&ots=43EBAqIkSJ&sig=npYylZteOmZAqv4f7SXAtJO9xr8&hl=en&sa=X&ved=0ahUKEwjFlf7-3p_cAhWPrVkKHXsFDWgQ6AEIUjAG#v=onepage&q=esterel%20external%20signal%20api&f=false
+Esterel relies on general-purpose languages such as C or C++ in two ways.       
+First, for portability, most Esterel compilers generate C instead of            
+assembly or some other executable representation.                               
+Such generated code provides an interface for passing events between the        
+environment and the running program.                                            
+Second, Esterel allows the use of data types and functions defined externally   
+(e.g., in C) to be used within an Esterel program.
+
+% Elm
+https://guide.elm-lang.org/interop/javascript.html
+Okay, in Elm, any communication with JavaScript goes through a port. Think of
+it like a hole in the side of your Elm program where you can send values in and
+out. These work exactly like the commands and subscriptions from the
+Architecture section. Sending values out to JS is a command. Listening for
+values coming in from JS is a subscription. Pretty neat!
+\end{comment}
+
+- excel define formulas for input but not the input itself
+
+output -> input
+
+the environment sits
+    inverse
+    reacts to output and generates inputs
+
+remain idle and react to the occurrence of events
 
@@
 %% Bibliography
-\bibliography{my}
+\bibliography{my,other}
 

### 2018-07-18 paper.tex
...

@@
 In reactive and event-based systems, execution is guided by an external
-environment which produces inputs and consumes output events to and from the
-applications.
+environment which generates inputs to the application and consumes outputs from
+it.
 Reactive languages provide dedicated syntax and semantics to deal with events
@@
 language.
-This separation exposes the environment as a rigid system component that
-evolves in separate from the application.
 %
@@
 targeting embedded systems in order to take control of the whole event loop,
-from the hardware input source and back to the hardware output sink.
+from a sensor input source and back to an actuator output.
 %
 We propose the new primitive in the context of the structured reactive language
-\CEU and discuss how it synergizes with the structured paradigm to provide
-automatic standby for applications and hot swapping for drivers during runtime.
+\CEU and discuss they synergize to provide automatic standby for applications
+and hot swapping of device drivers, while avoiding race conditions.
 \end{abstract}
@@
 Reactive applications interact continuously and in real time with the external
-world through input and output devices (e.g., buttons, displays, timers, etc.).
+world through sensors and actuators (e.g., buttons, displays, timers, etc.).
 %
@@
 while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
-out values to JavaScript as commands and listening for values as
+out values to JavaScript in commands and listening for values in
 subscriptions~\cite{frp.elm.ports}.
 
+%This separation exposes the environment as a rigid system component that
+%evolves in separate from the application.
 The event-based separation between the application and the environment is
-arguably inevitable and happens at the right level of abstraction, since it
-connects the application with the operating system resources through a
+arguably inevitable and also happens at the right level of abstraction, since
+it connects the application with the operating system resources through a
 non-invasive API.
 %
-However, it involves two languages and the programmer may have to unnecessarily
-deal with multiple syntaxes, incompatible type systems, and different address
-spaces.
+However, it involves two languages and the programmer may have to deal with
+multiple syntaxes, incompatible type systems, and different address spaces.
 %
 Furthermore, in the context of embedded systems, a proper host OS may even be
-absent or lacking enough drivers, which requires more low-level intervention
-from the application.
+absent or lacking enough device drivers, which requires more low-level
+intervention from the application.
 
@@
 for drivers during runtime.
+
+We present two uses cases that take advantage of the tight integration between
+the language and environment:
 

### 2018-07-27 paper.tex
...

@@
 %% use defaults for review submission.
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
+\acmConference[REBLS'18]{ACM SIGPLAN Workshop on Reactive and Event-Based Languages and Systems}{November 04, 2018}{Boston, MA, USA}
+\acmYear{2018}
 \acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
@@
 \newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
+
+\newcommand{\code}[1] {{\small{\texttt{#1}}}}
+
+\usepackage{listings}
+\lstdefinelanguage{ceu}{%
+  language=C,
+  morekeywords={%
+    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
+    await, bool, call, class, data, define, deterministic, do, each, else,
+    emit, end, escape, event, every, finalize, hor, implementation, in,
+    input, interface, loop, min, native, new, nohold, not, or, output, par,
+    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
+    code, public, private,
+    var, watching, when, with},
+}
+\lstset{
+  literate={~} {$\sim$}{1},
+  basicstyle=\footnotesize\ttfamily,
+  captionpos=b,
+  columns=flexible,
+  commentstyle=\footnotesize\itshape,
+  escapeinside={||},
+  frame=tb,
+  keepspaces=true,
+  keywordstyle=\bfseries,
+  language=ceu,
+  mathescape=true,
+  numbersep=4pt,
+  numberstyle=\tiny,
+  %xleftmargin=0.5cm,
+  %xrightmargin=0.5cm,
+  %upquote=true,
+}
+
+\usepackage{tikz}
+\usetikzlibrary{
+  arrows,
+  arrows.meta,
+  calc,
+  decorations.pathreplacing,
+  decorations.text,
+  math,
+  positioning,
+  shapes,
+}
 
@@
 %
-We propose the new primitive in the context of the structured reactive language
-\CEU and discuss they synergize to provide automatic standby for applications
-and hot swapping of device drivers, while avoiding race conditions.
+We propose the new asynchronous primitive in the context of the synchronous
+language \CEU and discuss how they synergize to avoid race conditions at
+compile time, support lexically-scoped drivers, and provide automatic standby
+for applications.
+%and hot swapping of device drivers
 \end{abstract}
@@
 %
-As illustrated in Figure~\ref{fig.loop}, devices can be encapsulated as a
+As illustrated in Figure~\ref{fig.env}, devices can be encapsulated as a
 single component, \emph{the environment}, which controls the system execution
@@
           events.
-\label{fig.loop}
+\label{fig.env}
 }
@@
 
-%This separation exposes the environment as a rigid system component that
-%evolves in separate from the application.
-The event-based separation between the application and the environment is
+The event-based interface between the application and the environment is
 arguably inevitable and also happens at the right level of abstraction, since
@@
 %
-However, it involves two languages and the programmer may have to deal with
-multiple syntaxes, incompatible type systems, and different address spaces.
+However, this separation exposes the environment as a rigid system component
+that evolves in separate from the application.
+%
+It also requires two languages and the programmer has to deal with multiple
+syntaxes, incompatible type systems, and different address spaces.
 %
@@
 
-In this work, we propose
-an interrupt handler primitive for a reactive language in
-order to take control of the whole event loop, from the
-hardware input source and back to the hardware output
-sink. We propose the new primitive in the context of
-the structured reactive language \CEU and discuss how
-it synergizes with the structured paradigm to provide
-automatic standby for applications and hot swapping
-for drivers during runtime.
+In this work, we propose interrupt service routines (ISRs) as an asynchronous
+primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
+ISRs empower applications to also implement their own device drivers and
+self-generate inputs, bypassing the need for a host environment.
+%
+\CEU targets resource-constrained architectures, such as Arduino-compatible
+microcontrollers, in which applications run in the bare metal, without
+operating system support.
+In this context, device drivers are typically libraries compiled together with
+the main program.
 
-We present two uses cases that take advantage of the tight integration between
-the language and environment:
+The lack of an OS opens an opportunity to explore a tighter integration between
+the application and its device drivers at the language level, resulting in an
+overall simplicity and tractability of the system.
+%
+In particular, \CEU is amenable to a simple static analysis to detect race
+conditions that we extended to encompass ISRs.
+%
+\CEU also provides a lexical finalization mechanism that we adopt in drivers to
+properly disable interrupts.
+%
+Finally, the synchronous semantics of \CEU enforces that applications react
+to inputs in bounded time and remain in an idle states susceptible to standby.
+With the help of drivers and a power management runtime, \CEU can put the
+microcontroller to sleep automatically at optimal sleeping modes after each
+input reaction.
 
-with different syntaxes and semantics.
-For instance, , type systems, and
+We implemented an extensible runtime that exposes hooks for the interrupts and
+the power manager, which we validated in two microcontrollers: an
+\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
+%
+We also implemented drivers for a variety of peripherals, such as GPIO, A/D
+converter, USART, SPI, and nRF24L01 transceiver.
+%
+Applications can share data buffers with drivers to avoid unnecessary copying,
+with the static guarantee that no data races will occur.
+%
+We demonstrate that applications built on top of these drivers show significant
+energy savings due to automatic standby.
 
-abstraction
-    - rigid
-    - two languages
-        - syntax and semantics
-            - integrates with features such as lexical scope, type system
-    - platforms w/o OS
-
-\newpage
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{sync-async}
-\caption{ Event loop in reactive systems.
-          The environment controls the application through input \& output
-          events.
-\label{fig.loop}
-}
-\end{figure}
+TODO: weakness
 
@@
 
+\section{\CEU: Structured Synchronous Reactive Programming}
+\label{sec.ceu}
+
+\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
+targeting resource-constrained embedded systems~\cite{ceu.sensys13}, with the
+characteristics that follow:
+%
+\begin{description}
+\item [Reactive:] code only executes in reactions to events and is idle most of
+    the time.
+\item [Structured:] programs use lexically-scoped structured control
+    mechanisms, such as \code{await} (to suspend a line of execution), and
+    \code{par} (to combine multiple lines of execution).
+\item [Synchronous:] reactions run atomically and to completion on each line of
+    execution, i.e., there's no implicit preemption or real parallelism.
+\end{description}
+
+Structured reactive programming lets developers write code in direct style,
+recovering from the inversion of control imposed by event-driven
+execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=h,
+  caption={A \CEU program that toggles the state of the LED whenever a radio
+           packet is received, terminating on a button press, always with the
+           LED off.},
+  label={lst.ceu},
+]
+output high/low LED;
+input  high/low BUTTON;
+input  Packet   RADIO_RECV;
+par/or do
+    var high/low v = await BUTTON (until v==low);
+with
+    finalize with
+        emit LED(low);
+    end
+    loop do
+        emit LED(high);
+        await RADIO_RECV;
+        emit LED(low);
+        await RADIO_RECV;
+    end
+end
+\end{lstlisting}
+
+The example in Listing~\ref{lst.ceu} illustrates the main characteristics of
+\CEU, namely event-driven I/O, lexically-scoped compositions and synchronous
+execution.
+%
+The program toggles the state of the LED whenever a radio packet is received,
+terminating on a button press, always with the LED off.
+%
+The program first declares the \code{LED} output, and the \code{BUTTON} and
+\code{RADIO\_RECV} input events (ln~1--3).
+The declarations include a payload type, i.e., each event occurrence carries
+a value of that type (\code{high/low} is a boolean type).
+%
+Then, the program uses a \code{par/or} composition to run two lines of
+execution, aka \emph{trails}, in parallel:
+a single-statement trail that waits for a button press before terminating
+(ln~5), and an endless loop that toggles the LED on and off whenever a radio
+packet is received (ln~10--15).
+Since we care about any reception occurrence, the example ignores the actual
+packet payload (ln~12,14).
+The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
+trail terminates, the LED will be unconditionally turned off~(ln~8).
+
+All communication between the application and the environment is done through
+the \code{await} and \code{emit} primitives, which awaits an input event and
+generates an output event, respectively.
+
+The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
+composition that terminates as soon as one of the trails terminates, but which
+also automatically finalizes the other trail(s).
+%
+In the example, when the button is pressed, not only the toggling loop will be
+aborted, but the \code{finalize} clause will turn the LED off unconditionally,
+since its enclosing block went out of scope.
+
+The synchronous execution model of \CEU dictates that reactions to input events
+are atomic and that incoming events are never lost.
+%
+The event loop in Figure~\ref{fig.env} indicates this behavior since the
+environment can only generate a new input after the application yields control
+and closes the loop.
+%
+In the example, even if two packets arrive simultaneously in the environment,
+the synchronous model provides the atomicity and responsiveness guarantees:
+    the first \code{await} will awake and turn the LED off atomically
+    (ln 12--13);
+    and the second \code{await} will awake in sequence (ln 14).
+%
+%Note that if multiple trails in parallel await the same input, they will all
+%awake from it during the same reaction.
+%In this case, \CEU dictates that each trail executes atomically, one after the
+%other, in lexical order.
+%This rule prevents race conditions for trails in parallel.
+
+Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
+%
+The continuous timeline in the environment shows an absolute reference clock
+with ``physical timestamps'' for the event occurrences (e.g., event~\code{C}
+occurs at 17ms521us).
+%
+The discrete timeline in the application shows how the same occurring events
+fit in the logical notion of time of \CEU.
+%
+The boot reaction \code{boot-0} happens before any input, at program startup.
+%
+Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
+is discrete, its corresponding reaction only executes afterwards, at logical
+instant~\code{A-1}.
+%
+Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
+postponed to execute at~\code{B-2}.
+%
+Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
+for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
+%
+Finally, event~\code{D} occurs during an idle period and can start immediately
+at \code{D-4}.
+%
+%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
+%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.
+
+\begingroup
+\begin{figure}[t]
+\centering
+\begin{tikzpicture}[x=1em,y=1em,
+  font=\footnotesize\sffamily,>=stealth,
+  time/.style={font=\scriptsize\sffamily},
+  cross/.style={inner sep=0pt,outer sep=0pt},
+  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
+  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
+    font=\ttfamily\footnotesize}]
+  \draw[->]
+  (0,0)
+    node[align=right,anchor=east,xshift=-.5em]
+        {Continuous\\Timeline\\(Env.)}
+    node[anchor=north west,xshift=-.5em,time]
+        {0ms000us}
+    node[anchor=north west,xshift=8.75em,time]
+        {17ms521us}
+  -- (18.5,0)
+    node[anchor=north east,xshift=.1em,time]
+        {30ms000us};
+  \node[cross] at (0,0){x};
+  %%
+  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
+  \coordinate(A) at (Ext-|4,0);
+  \coordinate(B) at (Ext-|5.5,0);
+  \coordinate(C) at (Ext-|9,0);
+  \coordinate(D) at (Ext-|15,0);
+  %%
+  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
+  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
+  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
+  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
+  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
+  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
+  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
+  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
+  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
+  %%
+  \draw[-]
+  (0,-5)
+    node[align=right,anchor=east,xshift=-.5em]
+        {Discrete\\Timeline\\(App.)}
+  -- (18.5,-5);
+  \node[cross](P0) at (0,-5){x};
+  \node[cross](P1) at (4.75,-5){x};
+  \node[cross](P2) at (10,-5){x};
+  \node[cross](P3) at (12.5,-5){x};
+  \node[cross](P4) at (14.5,-5){x};
+  \node[cross](P5) at (15.5,-5){x};
+  \node[cross](P6) at (18.5,-5){x};
+  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
+  \node[rng] at ($(P1)!.5!(P2)$){A-1};
+  \node[rng] at ($(P2)!.5!(P3)$){B-2};
+  \node[rng] at ($(P3)!.5!(P4)$){C-3};
+  \node[rng] at ($(P5)!.5!(P6)$){D-4};
+  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
+  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
+  %%
+  \begin{scope}[shorten >=1.5pt]
+  \draw[->] (Aevt) -- (P1);
+  \draw[->] (Bevt) -- (P2);
+  \draw[->] (Cevt) -- (P3);
+  \draw[->] (Devt) -- (P5);
+  \end{scope}
+\end{tikzpicture}
+\vskip-.6em
+%\includegraphics[width=\columnwidth]{tick_min}
+\caption{The discrete notion of time in \CEU.}
+\label{fig.ticks}
+\end{figure}
+\endgroup
+
+In order to guarantee responsiveness, the synchronous model relies on the
+\emph{synchronous hypothesis}, which states that reactions must be
+significantly faster than the rate of inputs.
+%
+For this reason, \CEU (like most synchronous languages) restricts itself and
+refuses unbounded loops at compile time.
+This guarantees that all reactions to the environment are computed in bounded
+time~\cite{ceu.lctes18}, ensuring that applications are always responsive to
+incoming events.
+
+\section{Interrupt Service Routines in \CEU}
+\label{sec.isrs}
+
+Interrupts service routines (ISRs) are software entry points that execute in
+response to hardware interrupts from peripherals such as timers and GPIOs.
+ISRs are at the lowest level interface between hardware and software and are
+the absolute source of inputs to programs.
+%
+Typically, an ISR starts to execute as soon as the hardware interrupt occurs
+and suspends the normal program flow abruptly.
+Such asynchronous behavior reflects the inherent concurrent nature of
+peripherals interacting with the external world.
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{sync-async}
+\caption{ An application in \CEU has a concurrent asynchronous side and a
+          predictable synchronous side that receives and reacts to inputs
+          atomically.
+\label{fig.async}
+}
+\end{figure}
+
+Asynchronous execution confronts the synchronous mindset of \CEU since the
+assumption that reactions are atomic no longer holds.
+Not only this may lead to race conditions at a fine grain, but may also affect
+the ordering of events at a coarse grain: the effect of an earlier event may be
+perceived after the effect of a later event.
+%
+Nevertheless, our goal is to preserve the well-behaved interaction between the
+\CEU application and the environment as of Figure~\ref{fig.env}, even in the
+presence of asynchronous ISRs as illustrated in Figure~\ref{fig.async}:
+%
+the original application is now represented by the synchronous \CEU and the
+environment by the asynchronous \CEU, all inside the same application.
+%
+This way, we want to push all subtleties of asynchronous execution into device
+drivers which emit input events to regular synchronous code in \CEU exactly
+as before.
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={XXX.},
+  label={lst.gpio.sync},
+  escapechar=\%,
+]
+output high/low PIN_13;
+input  high/low PIN_02;
+#include "gpio.ceu"
+
+emit PIN_13(low);
+loop do
+    var high/low v = await PIN_02;
+    emit PIN_13(v);
+end
+\end{lstlisting}
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={XXX.},
+  label={lst.gpio.async},
+  escapechar=\%,
+]
+// OUTPUT DRIVER
+
+{ pinMode(13, OUTPUT); }
+output (high/low v) PIN_13 do
+    { digitalWrite(13, @v); }
+end
+
+// INPUT DRIVER
+
+input high/low PIN_02;
+{
+    pinMode(2, INPUT_PULLUP);
+    EICRA = (EICRA & ~((1<<ISC00) | (1<<ISC01)))
+          | (CHANGE << ISC00);
+    EIMSK |= (1<<INT0);
+}
+spawn async/isr [INT0_vect] do
+    emit PIN_02({digitalRead(2)});
+end
+\end{lstlisting}
+
+Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} implements an application
+that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
+is on whenever the button is pressed.
+%
+The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
+the external world and includes the asynchronous side as a driver (ln 1--3).
+%
+The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
+input events.
+
+An \code{output} implementation (ln 4--6) is similar to a parameterized
+subroutine: whenever the application invokes \code{emit}, the output body
+executes atomically.
+\CEU supports inline C between curly braces (ln 3,5) with interpolation to
+evaluates \CEU expressions (e.g., \code{@v}).
+This allows drivers to take advantage of existing libraries in embedded
+toolchains, such as Arduino.
+%
+In the example, the driver sets pin 13 as output (ln 3) when it is included,
+and sets its state whenever it is emitted (ln 5).
+
+An \code{input} event implementation requires an ISR registered with the
+\code{spawn async/isr} primitive (ln 17--19), which is automatically invoked
+whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
+%
+An ISR in \CEU will always have an \code{emit} to an input event inside its
+body to awake the synchronous side.
+However, although the ISR executes asynchronously as soon as the interrupt
+occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
+an ongoing reaction on the synchronous side.
+%
+In the example, the input driver first configures pin 2 to behave as input and
+to generate an external interrupt on level transitions (ln 11--16).
+
+synchronous remains simple and well behaved
+with no low-level calls
+
+The application only communicates
+
+- finite is can be still much greater than 0 (order of milliseconds)
+    - which is a lot of time
+
 %% Acknowledgments

### 2018-08-01 paper.tex
...

@@
   language=ceu,
-  mathescape=true,
   numbersep=4pt,
@@
 \title{Where Do Events Come From?}
-\subtitle{Reactive Programming From The Ground Up}
+\subtitle{Reactive and Energy-Efficient Programming From The Ground Up}
 
@@
     the first \code{await} will awake and turn the LED off atomically
-    (ln 12--13);
-    and the second \code{await} will awake in sequence (ln 14).
+    (ln~12--13);
+    and the second \code{await} will awake in sequence (ln~14).
 %
@@
 
-\section{Interrupt Service Routines in \CEU}
+\subsection{Interrupt Service Routines}
 \label{sec.isrs}
@@
 
-Asynchronous execution confronts the synchronous mindset of \CEU since the
-assumption that reactions are atomic no longer holds.
+We propose to extend \CEU with asynchronous ISRs.
+However, asynchronous execution confronts the synchronous mindset of \CEU since
+the assumption that reactions are atomic no longer holds.
 Not only this may lead to race conditions at a fine grain, but may also affect
@@
   float=t,
-  caption={XXX.},
+  caption={Synchronous application that turns the LED on whenever the button
+           is pressed and off whenever it is unpressed.},
   label={lst.gpio.sync},
@@
 ]
-output high/low PIN_13;
-input  high/low PIN_02;
+output high/low PIN_13;     // connected to an LED
+input  high/low PIN_02;     // connected to a button
 #include "gpio.ceu"
@@
   float=t,
-  caption={XXX.},
+  caption={Asynchronous driver for GPIO which, on hardware interrupts, emits an
+           input event into a queue.},
   label={lst.gpio.async},
@@
     pinMode(2, INPUT_PULLUP);
-    EICRA = (EICRA & ~((1<<ISC00) | (1<<ISC01)))
-          | (CHANGE << ISC00);
-    EIMSK |= (1<<INT0);
+    EICRA |= (1 << ISC00);  // sets INT0
+    EIMSK |= (1 << INT0);   // turns on INT0
 }
@@
 that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
-is on whenever the button is pressed.
-%
+is on whenever the button is pressed and off whenever it is unpressed.
+
 The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
-the external world and includes the asynchronous side as a driver (ln 1--3).
-%
+the external world and includes the asynchronous side as a driver (ln~1--3).
+It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
+whenever the button changes (ln~7), changes the state of the LED with the new
+value (ln~8).
+
 The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
 input events.
-
-An \code{output} implementation (ln 4--6) is similar to a parameterized
+%
+An \code{output} implementation (ln~4--6) is similar to a parameterized
 subroutine: whenever the application invokes \code{emit}, the output body
 executes atomically.
-\CEU supports inline C between curly braces (ln 3,5) with interpolation to
+\CEU supports inline C between curly braces (ln~3,5) with interpolation to
 evaluates \CEU expressions (e.g., \code{@v}).
@@
 %
-In the example, the driver sets pin 13 as output (ln 3) when it is included,
-and sets its state whenever it is emitted (ln 5).
-
+In the example, the driver sets pin 13 as output (ln~3) when it is included,
+and sets its state whenever it is emitted (ln~5).
+%
 An \code{input} event implementation requires an ISR registered with the
-\code{spawn async/isr} primitive (ln 17--19), which is automatically invoked
+\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
 whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
 %
-An ISR in \CEU will always have an \code{emit} to an input event inside its
+An ISR in \CEU will typically have an \code{emit} to an input event inside its
 body to awake the synchronous side.
@@
 occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
-an ongoing reaction on the synchronous side.
+an ongoing reaction on the synchronous side (going into a queue to be
+discussed).
 %
 In the example, the input driver first configures pin 2 to behave as input and
-to generate an external interrupt on level transitions (ln 11--16).
+to generate an external interrupt on level transitions (ln~11--15).
 
-synchronous remains simple and well behaved
-with no low-level calls
+The example illustrates the clear separation between the application and the
+driver through an event-driven interface that reflects the desired architecture
+of Figures~\ref{fig.env}~and~\ref{fig.async}.
+%
+Now the whole application is written in \CEU:
+    the synchronous side remains well behaved with no low-level calls, and
+    the asynchronous side deals with all complexity in device drivers.
+%
+Nevertheless, drivers are typically write-once components developed by embedded
+specialists, which are reused in most applications.
+%
+Note that each architecture requires a mapping between the \code{async/isr} and
+the actual interrupt vector table, which we provide with our system.
 
-The application only communicates
+As we discuss in the next section, supporting ISRs in the same language and
+memory space of the application allows programmers to choose between efficiency
+and safety during development by providing (and experimenting with) multiple
+interfaces with different levels of abstraction.
+
+\subsection{Safe Shared-Memory Concurrency}
+\label{sec.ceu.shared}
+
+\begin{figure}[t!]
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[
+  numbers=right,
+  xrightmargin=3em,
+  basicstyle=\ttfamily\small,
+  label={lst.mem.a},
+  caption={Shared \code{x}},
+  escapechar=\%,
+]
+input void A;
+input void B;
+var int x = 1;
+par/and do
+    await A;
+    x = x + 1;
+with
+    await B;
+    x = x * 2;
+end
+\end{lstlisting}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+%\begin{lstlisting}
+\begin{lstlisting}[
+  xleftmargin=3em,
+  basicstyle=\ttfamily\small,
+  label={lst.mem.b},
+  caption={Shared \code{y}},
+  escapechar=\%,
+]
+input void A;
+// (empty line)
+var int y = 1;
+par/and do
+    await A;
+    y = y + 1;
+with
+    await A;
+    y = y * 2;
+end
+
+\end{lstlisting}
+\hskip1.5em
+\begin{minipage}[t]{.9\columnwidth}
+\end{minipage}%
+\end{minipage}%
+%\rule{8.4cm}{0.37pt}
+\caption{
+Accesses to shared \code{x} never concurrent.
+Accesses to shared \code{y} are concurrent but still deterministic.},
+\label{lst.mem}
+\end{figure}
+
+In \CEU, when multiple trails awake in the same reaction, they execute in
+lexical order, i.e., in the order they appear in the program source code.
+%
+In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
+assign to it in parallel trails (ln~6, 9).
+%
+In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
+reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
+occur simultaneously in accordance with the synchronous model.
+Hence, for the sequence of events \code{A->B}, \code{x} becomes
+\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
+\code{3}~(\code{(1*2)+1}).
+In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
+because they execute in reaction to the same event \code{A}.
+Nevertheless, because \CEU follows lexical order, the execution is still
+deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
+
+Regardless of deterministic execution, note that an apparently innocuous change
+in the order of trails may modify the behavior of programs.
+%
+To mitigate this threat, \CEU performs concurrency checks at compile time to
+detect conflicting accesses to shared variables:
+if a variable is written in a trail segment, then a concurrent trail segment
+cannot access that variable~\cite{ceu.sensys13}.
+%
+The static checks are optional (or raise warnings) and do not affect the
+semantics of the language.
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={The USART buffer is shared between the ISR and the application,
+           resulting in a pontential race condition.
+           \emph{(The driver and application code are in separate files.)}},
+  label={lst.usart},
+]
+// USART INTERFACE
+
+input none USART_RX;
+var[32*] byte rx_buf;             // `*' = circular
+
+// DRIVER
+
+spawn async/isr [USART_RX_vect] do
+    rx_buf = rx_buf .. [{UDR0}];
+    emit USART_RX;
+end
+
+// APPLICATION
+
+loop do
+    await USART_RX;
+    var int i;
+    loop i in [0 -> $rx_buf[ do   // `$' = length ($)
+        // uses rx_buf[i]
+    end
+    $rx_buf = 0;
+end
+\end{lstlisting}
+
+The addition of asynchronous ISRs poses real threats concerning race conditions
+in programs since interrupts occur at arbitrary points.
+%
+Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
+that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
+%
+The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
+and a circular buffer to avoid missing data (ln~4).
+The buffer is indispensable because the microcontroller cannot cope with the
+USART speed.
+%
+The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
+buffer (ln~9) and signal the application (ln~10).
+%
+As a possible race condition, note that the ISR may fire and append a new byte
+to the buffer (ln~9) just before the application clears it (ln~21), in which
+case the new byte will be lost forever.
+
+\CEU treats an \code{async/isr} as a block of code that runs in parallel with
+the rest of the program (hence the required prefix \code{spawn}).
+%
+This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
+Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
+a compile-time error.
+%
+The programmer is responsible to protect concurrent memory accesses with
+\code{atomic} blocks, which disables interrupts (supposedly) for a short period
+of time.
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={The USART driver now exposes a safer (and more friendly) interface
+           to the application.
+           \emph{(The driver and application code are in separate files.)}},
+  label={lst.usart.code},
+]
+// USART INTERFACE
+
+code Usart_RX (var&[] byte buf, var int n) -> none;
+
+// DRIVER
+
+input none USART_RX;
+var[32*] byte rx_buf;
+
+spawn async/isr [USART_RX_vect] do
+    rx_buf = rx_buf .. [{UDR0}];
+    emit USART_RX;
+end
+
+code Usart_RX (var&[] byte buf, var int n) -> none
+do
+    loop do
+        atomic do
+            buf = buf..rx_buf;
+            $rx_buf = 0;
+        end
+        if $buf >= n then
+            break;
+        end
+        await USART_RX;
+    end
+end
+
+// APPLICATION
+
+loop do
+    var[255] byte buf;
+    await Usart_RX(&buf, 10);
+    // uses buf
+end
+\end{lstlisting}
+
+Even so, we do not expect that application programmers are required to resolve
+race conditions.
+\CEU supports reactive abstractions (similar to coroutines) that help hiding
+the concurrency complexity inside the driver.
+%
+Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
+abstraction that expects a reference to a buffer and a number of bytes to
+receive.
+Now, the application (ln~29--35) invokes the abstraction (ln~33) with a local
+buffer (ln~32).
+%
+The driver (ln~5--27) now hides the low level interface of the previous example
+(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
+access to the shared buffer with an \code{atomic} block (ln~18--21).
+%
+The abstraction copies the driver buffer into the application buffer (ln~19)
+until the receiving the request number of bytes (ln~22).
+%
+On the one hand the extra copying incurs an overhead, but on the other hand it
+prevents race conditions with the ISR.
+
+\subsection{Energy-Efficient Runtime}
+\label{sec.energy}
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={XXX.},
+  label={lst.env},
+]
+evt_t queue[MAX];           // input queue
+void main () {
+  ceu_start();              // "boot reaction"
+  while (1) {
+    evt_t evt;
+    if (ceu_input(&evt)) {  // query input queue
+      ceu_sync(&evt);       // execute synchronous
+    } else {
+      ceu_pm_sleep();       // nothing to execute
+    }
+  }
+}
+\end{lstlisting}
+
+Our runtime of \CEU implements the event loop proposed in
+Figure~\ref{fig.async}, which alternates between synchronous, well-behaved
+execution, and asynchronous ISRs.
+%
+Since the language now has full control over the event loop and only executes
+from interrupts, we want the runtime to enter in sleep mode automatically to
+save energy.
+
+Listing~\ref{lst.env} is a realization of this architecture:
+%
+The runtime defines an input queue (ln~1) in which ISRs enqueue new events
+(e.g., Listing~\ref{lst.usart.code}, ln~12).
+%
+In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
+which starts the program to spawn the ISRs and reach the first \code{await}
+statements in the multiple lines of code.
+%
+Then, the runtime enters an infinite loop (ln~4--11) that queries the input
+queue (ln~6) to awake the program (ln~7).
+%
+Note that ISRs execute asynchronously with the loop, but they typically only
+emit an input event into the queue which will be queried in a subsequent
+iteration.
+%
+Note also that if the queue is empty, the runtime enters in sleep mode to save
+energy (ln~9), and will only awake on a new hardware interrupt.
+
+Microcontrollers typically support multiple levels of sleep mode, each
+progressively saves more energy at the expense of keeping less functionality
+active.
+%
+As an example, the least efficient mode of the \emph{ATmega328P} allows for
+timer interrupts since it keeps its internal clock active, while the most
+efficient mode turns off all peripherals and can only awake from external
+interrupts.
+
+Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
+(ln~9):
+%
+The first configuration is a \emph{nop} which simply keeps the event loop
+running all the time without ever sleeping.
+This configuration is useful when introducing new platforms.
+%
+The second configuration chooses a (inefficient) sleep mode that keeps all
+peripherals active.
+Although this configuration is not the most energy efficient, at least, it
+requires no extra assistance from the drivers.
+%
+The third configuration keeps a bit vector of the active drivers during runtime
+and chooses the most efficient mode, querying the vector every time
+\code{ceu\_pm\_sleep} is called.
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={XXX.
+           \emph{(The power manager, driver and application code are in separate files.)}},
+  label={lst.pm},
+  escapechar=\%,
+]
+// POWER MANAGER (in C)
+
+enum {
+    CEU_PM_ADC,
+    CEU_PM_TIMER1,
+    CEU_PM_USART,
+    ...
+};
+
+void ceu_pm_sleep (void) {
+    if (ceu_pm_get(CEU_PM_USART) || ...) {
+        sleep_mode_1(...);
+    } else if (ceu_pm_get(CEU_PM_ADC)) {
+        sleep_mode_2(...);
+    } else {
+        sleep_mode_3(...);
+    }
+}
+
+// USART DRIVER (in Ceu)
+
+code Usart_RX (var&[] byte buf, var int n) -> none
+do
+    {ceu_pm_set(CEU_PM_USART, 1);}
+    do finalize with
+        {ceu_pm_set(CEU_PM_USART, 0);}
+    end
+    ...   // same as in Listing 7
+end
+
+// APPLICATION (in Ceu)
+
+input high/low PIN_02;      // connected to a button
+loop do
+    par/or do
+        await PIN_02;
+    with
+        var[255] byte buf;
+        await Usart_RX(&buf, 10);
+        // uses buf
+    end
+    await PIN_02;
+end
+\end{lstlisting}
+
+The third configuration achieves optimal efficiency but requires a tight
+interaction between the drivers and the power management runtime.
+%
+Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
+driver (ln~20--29), and an illustrative application (ln~31--43).
+%
+The application is a loop that awaits in parallel for a button press (ln~36) or
+receiving 10 bytes (ln~39).
+Let's call this initial state \code{STATE-A}.
+The \code{par/or} terminates when either of them occurs, going to the next line
+that awaits another button click (ln~42).
+Let's call this other state \code{STATE-B}.
+After the button click, the program loops back to \code{STATE-A}.
+%
+While in \code{STATE-A}, the program can awake from USART and external
+interrupts, which means that the power manager should choose a sleep mode in
+which the USART remains operational.
+%
+While in \code{STATE-B}, only external interrupts should awake the program,
+which menas that the power manager may use the most efficient sleep mode.
+%
+The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
+before sleeping, queries their states (ln~11,13) to choose the most appropriate
+sleep mode (ln~12,14,16).
+%
+The USART driver needs to be extended with calls to enable and disable the
+USART state inside the power manager: just before awaiting, the driver enables
+the USART (ln~24) and registers a finalization block to unregister it on
+termination (ln~25--27).
+Termination may occur either directly after receiving the requested number of
+bytes, or indirectly if the \code{par/or} in the application terminates on a
+button click (ln~36).
+
+Note that applications need not to be explicit about energy management to take
+advantage of sleep modes.
+All happens automatically because of the synchronous-reactive execution model
+and energy-aware runtime of \CEU.
+
+\section{Related Work}
+
+- TinyOS
+    - async, race condition avoidance
+        - nesc manual
+    - ICEM, standby
+        - icem paper
+
+- CRP
+
+- finalize
+- structured programming
+    - lexical scoping vs manual
+    - no turn off
+
+- citar ICEM o mais cedo nas duas ocasioes (async, energy)
+- citar TOS/ICEM na introducao como maior influencia
+    - ja no abstract?
+
+- falar que async/isr nao contem nada de Ceu, apenas shared memory e nocao de estar em paralelo
+    - compartilhamento de memoria
+        - velocidade
+    - ajuda do compilador p/ detectar acesso concorrente
 

### 2018-08-02 paper.tex
...

@@
 In this work, we propose an interrupt handler primitive for a reactive language
-targeting embedded systems in order to take control of the whole event loop,
+targeting embedded systems in order to take control of the whole event loop:
 from a sensor input source and back to an actuator output.
@@
 Reactive applications interact continuously and in real time with the external
-world through sensors and actuators (e.g., buttons, displays, timers, etc.).
+world through sensors and actuators peripherals (e.g., buttons, displays,
+timers, etc.).
 %
 These interactions are typically represented as input events flowing from the
-devices to the application and as output events flowing from the application to
-the devices.
-%
-As illustrated in Figure~\ref{fig.env}, devices can be encapsulated as a
-single component, \emph{the environment}, which controls the system execution
+peripherals to the application and as output events flowing from the
+application to the peripherals.
+%
+As illustrated in Figure~\ref{fig.env}, all peripherals can be encapsulated as
+a single component, \emph{the environment}, which controls the system execution
 in an event loop:
@@
 The event-based interface between the application and the environment is
-arguably inevitable and also happens at the right level of abstraction, since
-it connects the application with the operating system resources through a
+arguably inevitable and also happens at the appropriate layer, since it
+connects the application with the operating system resources through a
 non-invasive API.
@@
 primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
-ISRs empower applications to also implement their own device drivers and
-self-generate inputs, bypassing the need for a host environment.
+ISRs empower reactive applications to also implement their own device drivers
+and self-generate inputs, bypassing the need for a host environment.
 %
@@
 %
-In particular, \CEU is amenable to a simple static analysis to detect race
-conditions that we extended to encompass ISRs.
+In particular, \CEU is already amenable to a simple static analysis to detect
+race conditions which we extended to also encompass ISRs.
+Applications can share data buffers with drivers to avoid unnecessary copying,
+with the static guarantee that no data races will occur.
 %
 \CEU also provides a lexical finalization mechanism that we adopt in drivers to
-properly disable interrupts.
+properly disable interrupts and turn off peripherals completely.
+For instance, we can delimit the scope of a driver similarly to a
+lexically-scoped variables.
 %
@@
 to inputs in bounded time and remain in an idle states susceptible to standby.
-With the help of drivers and a power management runtime, \CEU can put the
-microcontroller to sleep automatically at optimal sleeping modes after each
-input reaction.
+With the help of drivers and a power manager, we can put the microcontroller to
+sleep automatically at optimal sleeping modes after each input reaction.
 
@@
 %
-Applications can share data buffers with drivers to avoid unnecessary copying,
-with the static guarantee that no data races will occur.
-%
-We demonstrate that applications built on top of these drivers show significant
-energy savings due to automatic standby.
+The applications built on top of these drivers show significant energy savings
+due to automatic standby.
 
 TODO: weakness
-
-what would be
-
-output (int,int) O;
-
-becomes
-
-output (int x, int y) O do
-end
 
@@
 
-\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
-targeting resource-constrained embedded systems~\cite{ceu.sensys13}, with the
+\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{ceu.tecs17} reactive
+programming language targeting resource-constrained embedded systems, with the
 characteristics that follow:
@@
 \item [Structured:] programs use lexically-scoped structured control
-    mechanisms, such as \code{await} (to suspend a line of execution), and
-    \code{par} (to combine multiple lines of execution).
+    mechanisms such as \code{spawn} and \code{await} (to create and suspend
+    lines of execution).
 \item [Synchronous:] reactions run atomically and to completion on each line of
@@
 
-The example in Listing~\ref{lst.ceu} illustrates the main characteristics of
-\CEU, namely event-driven I/O, lexically-scoped compositions and synchronous
-execution.
+Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
+event-driven I/O, lexically-scoped compositions, and synchronous execution.
 %
@@
 packet is received (ln~10--15).
-Since we care about any reception occurrence, the example ignores the actual
-packet payload (ln~12,14).
 The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
-trail terminates, the LED will be unconditionally turned off~(ln~8).
+trail terminates (e.g., from a button press), the LED will be unconditionally
+turned off~(ln~8).
 
@@
 aborted, but the \code{finalize} clause will turn the LED off unconditionally,
-since its enclosing block went out of scope.
+since its enclosing block goes out of scope.
 
@@
 %
-The event loop in Figure~\ref{fig.env} indicates this behavior since the
+The event loop in Figure~\ref{fig.env} suggests this behavior since the
 environment can only generate a new input after the application yields control
@@
 In the example, even if two packets arrive simultaneously in the environment,
-the synchronous model provides the atomicity and responsiveness guarantees:
-    the first \code{await} will awake and turn the LED off atomically
+the synchronous model ensures atomicity and responsiveness:
+    the first \code{await} awakes and turns the LED off atomically
     (ln~12--13);
@@
 %
-The continuous timeline in the environment shows an absolute reference clock
-with ``physical timestamps'' for the event occurrences (e.g., event~\code{C}
-occurs at 17ms521us).
+The continuous timeline (representing the environment) shows
+``physical timestamps'' for the event occurrences (e.g., event~\code{C}
+occurs exactly at 17ms521us).
 %
@@
 %
-The boot reaction \code{boot-0} happens before any input, at program startup.
+The boot reaction \code{boot-0} happens before any input and starts the
+program.
 %
 Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
-is discrete, its corresponding reaction only executes afterwards, at logical
+is discrete, its corresponding reaction only executes at logical
 instant~\code{A-1}.
@@
 Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
-for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
+for~\code{B-2} to execute at~\code{C-3}.
 %
@@
 
+- TODO adc
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=t,
+  caption={XXX.},
+  label={lst.gpio.async},
+  escapechar=\%,
+]
+// A/D DRIVER (adc.ceu)
+
+output int ADC_REQUEST;
+input  int ADC_DONE;
+
+do finalize with
+  {
+    ADCSRA &= B01111111;  // disables A/D converter
+    ACSR    = B10000000;  // disables comparator
+    DIDR0  |= B00111111;  // disables pins
+  }
+end
+
+...     // driver implementation
+
+// APPLICATION
+
+loop do
+    var int v;
+    do
+        #include "adc.ceu"
+        emit ADC_REQUEST(A0);
+        v = await ADC_DONE;
+    end
+    ...     // uses v
+    await 1h;
+end
+\end{lstlisting}
+
 \subsection{Safe Shared-Memory Concurrency}

### 2018-08-02 paper.tex
...

@@
 response to hardware interrupts from peripherals such as timers and GPIOs.
-ISRs are at the lowest level interface between hardware and software and are
-the absolute source of inputs to programs.
-%
-Typically, an ISR starts to execute as soon as the hardware interrupt occurs
-and suspends the normal program flow abruptly.
+ISRs are at the lowest interface layer between the hardware and software and
+are the absolute source of inputs to programs.
+%
+An ISR starts to execute as soon as the hardware interrupt occurs, suspending
+the ongoing program flow abruptly.
 Such asynchronous behavior reflects the inherent concurrent nature of
@@
 %
-Nevertheless, our goal is to preserve the well-behaved interaction between the
-\CEU application and the environment as of Figure~\ref{fig.env}, even in the
-presence of asynchronous ISRs as illustrated in Figure~\ref{fig.async}:
-%
-the original application is now represented by the synchronous \CEU and the
-environment by the asynchronous \CEU, all inside the same application.
-%
-This way, we want to push all subtleties of asynchronous execution into device
+Still, our goal is to preserve the well-behaved interaction between the \CEU
+application and the environment even in the presence of asynchronous ISRs.
+Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env} to fit all
+inside the same \CEU application, now with the original application represented
+as the synchronous side and the environment as the asynchronous side.
+%
+Our idea is to push all subtleties of asynchronous execution into device
 drivers which emit input events to regular synchronous code in \CEU exactly
@@
 
-Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} implements an application
-that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
-is on whenever the button is pressed and off whenever it is unpressed.
-
-The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
-the external world and includes the asynchronous side as a driver (ln~1--3).
+As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
+GPIOs to virtually connect a button (pin 02) to an LED (pin 13) such that the
+LED is on whenever the button is pressed and off whenever it is unpressed.
+
+Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
+with the external world and includes the asynchronous side as a driver
+(ln~1--3).
 It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
@@
 
-The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
+Listing~\ref{lst.gpio.async}, the asynchronous side, implements the output and
 input events.
@@
 executes atomically.
-\CEU supports inline C between curly braces (ln~3,5) with interpolation to
-evaluates \CEU expressions (e.g., \code{@v}).
+\CEU supports inline C between curly braces (ln~3,5) and interpolation to
+evaluate \CEU expressions (e.g., \code{@v}).
 This allows drivers to take advantage of existing libraries in embedded
@@
 %
-In the example, the driver sets pin 13 as output (ln~3) when it is included,
-and sets its state whenever it is emitted (ln~5).
-%
-An \code{input} event implementation requires an ISR registered with the
+In the example, when the driver is included, it configures pin 13 as output
+(ln~3) and sets its state whenever \code{PIN\_13} is emitted (ln~5).
+%
+An \code{input} event implementation uses an ISR registered with the
 \code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
@@
 %
-An ISR in \CEU will typically have an \code{emit} to an input event inside its
-body to awake the synchronous side.
-However, although the ISR executes asynchronously as soon as the interrupt
-occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
-an ongoing reaction on the synchronous side (going into a queue to be
-discussed).
-%
-In the example, the input driver first configures pin 2 to behave as input and
-to generate an external interrupt on level transitions (ln~11--15).
+An ISR in \CEU will typically emit an input event inside its body to awake the
+synchronous side (ln~17).
+However, although the ISR executes asynchronously when the interrupt occurs,
+the \code{emit} goes into a queue (to be discussed) and does not interrupt the
+ongoing (suspended) reaction on the synchronous side.
+%
+In the example, the input driver also configures pin 2 to behave as input and
+to generate external interrupts on level transitions (ln~11--15).
 
@@
 %
-Now the whole application is written in \CEU:
+Now, the whole application is written in \CEU:
     the synchronous side remains well behaved with no low-level calls, and
-    the asynchronous side deals with all complexity in device drivers.
+    the asynchronous side deals with all complexity of device drivers.
 %
@@
 memory space of the application allows programmers to choose between efficiency
-and safety during development by providing (and experimenting with) multiple
-interfaces with different levels of abstraction.
-
-- TODO adc
+and safety during development by providing multiple interfaces with different
+levels of abstraction.
 
@@
   caption={XXX.},
-  label={lst.gpio.async},
+  label={lst.adc},
   escapechar=\%,
@@
   end
-
-...     // driver implementation
-
-// APPLICATION
+  ...     // driver initialization
+  ...     // input / output implementations
+
+// APPLICATION (main.ceu)
 
@@
     do
-        #include "adc.ceu"
+      #include "adc.ceu" // driver contents above
       emit ADC_REQUEST(A0);
@@
     end
-    ...     // uses v
+    ...                  // uses sensor value "v"
     await 1h;
@@
 \end{lstlisting}
+
+The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
+driver for the A/D converter.
+Note that the driver (ln~1--14) and application (ln~16--27) are actually in
+separate files.
+%
+The application is a typical periodic sensor sampling loop, but which
+designates an explicit \code{do--end} scope to execute the driver (ln~21) and
+use it (ln~22--23).
+%
+The driver specifies a finalization code (ln~6--12) which executes
+unconditionally whenever it goes out of scope (ln~24).
+This code completely disables all A/D functionality, saving energy.
+%
+As discussed for Listing~\ref{lst.ceu}, finalizers also execute due to
+\code{par/or} termination (or other abortion mechanisms of \CEU).
+%
+Note that the driver interface (ln~3--4) is visible only inside the block and
+thus cannot be inadvertently used outside it.
 

### 2018-08-03 paper.tex
...

@@
 \end{comment}
-
-- excel define formulas for input but not the input itself
-
-output -> input
-
-the environment sits
-    inverse
-    reacts to output and generates inputs
-
-remain idle and react to the occurrence of events
 
@@
 composition that terminates as soon as one of the trails terminates, but which
-also automatically finalizes the other trail(s).
+also automatically aborts and finalizes the other trail(s).
 %
@@
 since its enclosing block goes out of scope.
+%
+The \code{par/or} is regarded as an \emph{orthogonal preemption
+primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
+in order to affect each other.
 
@@
 
-\subsection{Interrupt Service Routines}
+\subsection{Lexically-Scoped Interrupt Service Routines}
 \label{sec.isrs}
@@
 unconditionally whenever it goes out of scope (ln~24).
-This code completely disables all A/D functionality, saving energy.
-%
-As discussed for Listing~\ref{lst.ceu}, finalizers also execute due to
-\code{par/or} termination (or other abortion mechanisms of \CEU).
+This code completely disables all A/D functionality to save energy.
+%
+As discussed for Listing~\ref{lst.ceu}, the finalizer would also sensibly
+execute due to a \code{par/or} termination enclosing the driver.
 %

### 2018-08-03 paper.tex
...

@@
 In reactive and event-based systems, execution is guided by an external
-environment which generates inputs to the application and consumes outputs from
+environment that generates inputs to the application and consumes outputs from
 it.
@@
 %% comma separated list
-\keywords{interrupt service routine,
-          sleep mode,
-          synchronous reactive programming}
+\keywords{
+    interrupt service routines,
+    reactive programming,
+    sleep modes,
+    standby,
+    synchronous/asynchronous execution
+}
 
@@
 Reactive applications interact continuously and in real time with the external
-world through sensors and actuators peripherals (e.g., buttons, displays,
-timers, etc.).
-%
-These interactions are typically represented as input events flowing from the
-peripherals to the application and as output events flowing from the
-application to the peripherals.
-%
-As illustrated in Figure~\ref{fig.env}, all peripherals can be encapsulated as
-a single component, \emph{the environment}, which controls the system execution
+world through hardware peripherals such as sensors and actuators (e.g.,
+buttons, displays, timers, etc.).
+%
+These interactions are typically represented in software as input events
+flowing from the peripherals to the application and as output events flowing
+from the application to the peripherals.
+%
+As illustrated in Figure~\ref{fig.env}, peripherals can be encapsulated as a
+single component, \emph{the environment}, which connects with the application
 in an event loop:
-the application sits idle waiting for an input;
-the environment awakes the application on the occurrence of an input;
-the application reacts to the input and generates back one or more outputs;
+the application sits idle until the environment awakes it on the occurrence of
+an input;
+the application reacts to the input and generates back one or more outputs
+which affect the environment;
 the application becomes idle and the loop restarts.
@@
 The environment is typically implemented in a host language (e.g., C) and
-controls the main event loop, invoking entry points in the reactive language
-runtime on the occurrence of inputs and also receiving output calls, both
-through a documented API.
+controls the main event loop, invoking entry points into the reactive language
+runtime on the occurrence of inputs, and also receiving output calls from it,
+both through a documented API.
 %
@@
 while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
-out values to JavaScript in commands and listening for values in
+out values to JavaScript as commands and listening for values as
 subscriptions~\cite{frp.elm.ports}.
@@
 %
-However, this separation exposes the environment as a rigid system component
+However, this separation reveals the environment as a rigid system component
 that evolves in separate from the application.
@@
 In particular, \CEU is already amenable to a simple static analysis to detect
-race conditions which we extended to also encompass ISRs.
+race conditions which we extended in this work to also encompass ISRs.
 Applications can share data buffers with drivers to avoid unnecessary copying,
-with the static guarantee that no data races will occur.
+with some static guarantee that no data races will occur.
 %
@@
 Finally, the synchronous semantics of \CEU enforces that applications react
-to inputs in bounded time and remain in an idle states susceptible to standby.
+to inputs in bounded time and remain in idle states susceptible to standby.
 With the help of drivers and a power manager, we can put the microcontroller to
@@
 We implemented an extensible runtime that exposes hooks for the interrupts and
-the power manager, which we validated in two microcontrollers: an
+power manager, which we validated in two microcontrollers: an
 \emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
@@
 par/or do
-    var high/low v = await BUTTON (until v==low);
+    var high/low v = await BUTTON until (v==low);
 with
@@
 %
-Then, the program uses a \code{par/or} composition to run two lines of
-execution, aka \emph{trails}, in parallel:
+Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
+of execution, aka \emph{trails}, in parallel:
 a single-statement trail that waits for a button press before terminating
@@
 %
-In the example, when the button is pressed, not only the toggling loop will be
-aborted, but the \code{finalize} clause will turn the LED off unconditionally,
-since its enclosing block goes out of scope.
+In the example, when the button is pressed (ln~5), not only the toggling loop
+will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
+off unconditionally (ln~8), since its enclosing block goes out of scope.
 %
@@
 %
-In the example, even if two packets arrive simultaneously in the environment,
-the synchronous model ensures atomicity and responsiveness:
+In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
+environment, the synchronous model ensures atomicity and responsiveness in the
+application:
     the first \code{await} awakes and turns the LED off atomically
@@
 %
-For this reason, \CEU (like most synchronous languages) restricts itself and
-refuses unbounded loops at compile time.
-This guarantees that all reactions to the environment are computed in bounded
-time~\cite{ceu.lctes18}, ensuring that applications are always responsive to
-incoming events.
+For this reason, \CEU (like most synchronous languages) refuses unbounded loops
+at compile time.
+This restriction guarantees that all reactions to the environment are computed
+in bounded time~\cite{ceu.lctes18}, ensuring that applications are always
+responsive to upcoming events.
 
@@
 application and the environment even in the presence of asynchronous ISRs.
-Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env} to fit all
-inside the same \CEU application, now with the original application represented
-as the synchronous side and the environment as the asynchronous side.
+Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
+original application represented as the synchronous side and the environment as
+the asynchronous side, all inside the same \CEU application.
 %
 Our idea is to push all subtleties of asynchronous execution into device
-drivers which emit input events to regular synchronous code in \CEU exactly
+drivers, which emit input events to regular synchronous code in \CEU, exactly
 as before.
@@
 input  high/low PIN_02;  // connected to a button
-#include "gpio.ceu"
+#include "gpio.ceu"      // GPIO driver
 
@@
 As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
-GPIOs to virtually connect a button (pin 02) to an LED (pin 13) such that the
-LED is on whenever the button is pressed and off whenever it is unpressed.
+GPIOs to connect a button (pin 02) to an LED (pin 13) via software such that
+the LED is on whenever the button is pressed and off whenever it is unpressed.
 
@@
 (ln~1--3).
-It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
-whenever the button changes (ln~7), changes the state of the LED with the new
+It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
+whenever the button changes (ln~7), toggles the state of the LED with the new
 value (ln~8).
 
-Listing~\ref{lst.gpio.async}, the asynchronous side, implements the output and
-input events.
+Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
+the output and input events.
 %
@@
 executes atomically.
-\CEU supports inline C between curly braces (ln~3,5) and interpolation to
+\CEU supports inline C between curly braces (ln~3,5) with interpolation to
 evaluate \CEU expressions (e.g., \code{@v}).
-This allows drivers to take advantage of existing libraries in embedded
+This allows drivers to take advantage of existing libraries of embedded
 toolchains, such as Arduino.
@@
 In the example, when the driver is included, it configures pin 13 as output
-(ln~3) and sets its state whenever \code{PIN\_13} is emitted (ln~5).
+(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
 %
@@
     the synchronous side remains well behaved with no low-level calls, and
-    the asynchronous side deals with all complexity of device drivers.
+    the asynchronous side deals with the complexity of device drivers.
 %
@@
 %
-Note that each architecture requires a mapping between the \code{async/isr} and
-the actual interrupt vector table, which we provide with our system.
-
-As we discuss in the next section, supporting ISRs in the same language and
-memory space of the application allows programmers to choose between efficiency
-and safety during development by providing multiple interfaces with different
-levels of abstraction.
+Note that each supported architecture requires a mapping between the
+\code{async/isr} and the actual interrupt vector table (which we provide
+for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).
 
@@
   float=t,
-  caption={XXX.},
+  caption={The A/D driver (ln~1--14) is included in the application inside a
+           lexically-scoped block (ln~20--24), which turns off all driver
+           functionalities automatically on termination.},
   label={lst.adc},
@@
 The application is a typical periodic sensor sampling loop, but which
-designates an explicit \code{do--end} scope to execute the driver (ln~21) and
-use it (ln~22--23).
+designates an explicit \code{do-end} block (ln~20--24) to include the driver
+(ln~21) and use it (ln~22--23).
 %
 The driver specifies a finalization code (ln~6--12) which executes
-unconditionally whenever it goes out of scope (ln~24).
-This code completely disables all A/D functionality to save energy.
-%
-As discussed for Listing~\ref{lst.ceu}, the finalizer would also sensibly
-execute due to a \code{par/or} termination enclosing the driver.
+unconditionally whenever its enclosing block goes out of scope (ln~24).
+The finalize completely disables all A/D functionality to save energy.
+%
+As discussed in the previous section, the finalizer would also execute if
+aborted from a \code{par/or} enclosing the driver.
 %
@@
 In \CEU, when multiple trails awake in the same reaction, they execute in
-lexical order, i.e., in the order they appear in the program source code.
+lexical order, i.e., in the order they appear in the source code.
 %
@@
 occur simultaneously in accordance with the synchronous model.
-Hence, for the sequence of events \code{A->B}, \code{x} becomes
-\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
-\code{3}~(\code{(1*2)+1}).
+%Hence, for the sequence of events \code{A->B}, \code{x} becomes
+%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
+%\code{3}~(\code{(1*2)+1}).
 In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
@@
 deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
-
-Regardless of deterministic execution, note that an apparently innocuous change
-in the order of trails may modify the behavior of programs.
-%
-To mitigate this threat, \CEU performs concurrency checks at compile time to
-detect conflicting accesses to shared variables:
+%
+Even so, \CEU performs (optional) concurrency checks at compile time to detect
+conflicting accesses to shared variables:
 if a variable is written in a trail segment, then a concurrent trail segment
@@
 %
-The static checks are optional (or raise warnings) and do not affect the
-semantics of the language.
+%The static checks are optional (or raise warnings) and do not affect the
+%semantics of the language.
 
@@
 The addition of asynchronous ISRs poses real threats concerning race conditions
-in programs since interrupts occur at arbitrary points.
+since they interrupt programs at arbitrary points.
 %
@@
 The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
-and a circular buffer to avoid missing data (ln~4).
-The buffer is indispensable because the microcontroller cannot cope with the
-USART speed.
+and a circular buffer to prevent data loss (ln~4).
+The buffer is indispensable because the microcontroller may cope with the USART 
+speed.
 %
 The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
-buffer (ln~9) and signal the application (ln~10).
+buffer (ln~9) and signals the application (ln~10).
 %
@@
 
-Even so, we do not expect that application programmers are required to resolve
-race conditions.
+However, we do not expect that application programmers should be required to
+resolve race conditions.
 \CEU supports reactive abstractions (similar to coroutines) that help hiding
@@
 receive.
-Now, the application (ln~29--35) invokes the abstraction (ln~33) with a local
-buffer (ln~32).
+Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
+local buffer (ln~32).
 %
@@
 The abstraction copies the driver buffer into the application buffer (ln~19)
-until the receiving the request number of bytes (ln~22).
-%
-On the one hand the extra copying incurs an overhead, but on the other hand it
-prevents race conditions with the ISR.
+up to the requested number of bytes (ln~22).
+%
+On the one hand the extra copying incurs a memory and runtime overhead, but on
+the other hand it prevents race conditions with the ISR.
+
+Support for ISRs in the same language and memory space of the application
+allows programmers to choose between efficiency and safety during development
+by providing multiple interfaces with different levels of abstraction.
 
@@
   float=t,
-  caption={XXX.},
+  caption={Overall runtime architecture of \CEU with an input queue (ln~1),
+           event loop (ln~4--11), synchronous execution (ln~7), and standby
+           mode (ln~9).},
   label={lst.env},
@@
 
-Our runtime of \CEU implements the event loop proposed in
-Figure~\ref{fig.async}, which alternates between synchronous, well-behaved
-execution, and asynchronous ISRs.
+Our runtime implements the event loop proposed in Figure~\ref{fig.async}, which
+alternates between synchronous, well-behaved execution, and asynchronous,
+unpredictable ISRs.
 %
 Since the language now has full control over the event loop and only executes
-from interrupts, we want the runtime to enter in sleep mode automatically to
-save energy.
+from interrupts, it is possible for the runtime to enter in sleep mode
+automatically to save energy.
 
@@
 In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
-which starts the program to spawn the ISRs and reach the first \code{await}
+which starts the program, spawns the ISRs, and reaches the first \code{await}
 statements in the multiple lines of code.
@@
 The third configuration keeps a bit vector of the active drivers during runtime
-and chooses the most efficient mode, querying the vector every time
+and chooses the most efficient mode, querying this vector every time
 \code{ceu\_pm\_sleep} is called.
@@
   float=t,
-  caption={XXX.
-           \emph{(The power manager, driver and application code are in separate files.)}},
+  caption={Interaction between the power manager, USART driver, and application
+           (each code is actually in a separate file).},
   label={lst.pm},
@@
 While in \code{STATE-B}, only external interrupts should awake the program,
-which menas that the power manager may use the most efficient sleep mode.
+which means that the power manager may use the most efficient sleep mode.
 %
@@
 USART state inside the power manager: just before awaiting, the driver enables
-the USART (ln~24) and registers a finalization block to unregister it on
+the USART (ln~24) and creates a finalization block to unregister it on
 termination (ln~25--27).

### 2018-08-04 paper.tex
...

@@
 
-TODO: weakness
+Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
+wireless devices, which also provides asynchronous events with a compile-time
+data race detector~\cite{wsn.nesc}, and automatic energy management triggered
+from idle CPU states~\cite{wsn.icem}.
+%
+We adapted these ideas to the structured reactive programming model of \CEU,
+and refined them with stronger guarantees due to lexical scope and automatic
+finalization.
 
@@
 
+TODO: energy savings
+
 \section{Related Work}
 
-- TinyOS
-    - async, race condition avoidance
-        - nesc manual
-    - ICEM, standby
-        - icem paper
+Our work is largely inspired by TinyOS~\cite{wsn.tos} and its companion
+language nesC~\cite{wsn.nesc}, which is a callback-driven dialect of C.
+%
+By default, callbacks are synchronous but can be annotated as \code{async} when
+triggered directly from interrupts.
+In this case, a static data-race detector ensures that concurrent accesses are
+protected by atomic sections.
+%
+Since the TinyOS runtime is also based on an event loop, it can easily detect
+idle periods and put the CPU to sleep.
+Device drivers also cooperate with the runtime to designate the best sleep
+mode~\cite{wsn.icem}.
+
+In comparison to nesC, our proposal takes advantage of the structured
+mechanisms of \CEU to offer even stronger guarantees.
+%
+For instance, nesC cannot provide a safe abortion mechanism, since the notion
+of control flow is entirely dynamic, typically encoded inside ordinary global
+variables that represent state machines.
+%
+For this reason, some manual bookkeeping is required to avoid that resources
+eventually leak (e.g., memory or energy).
+As an example, if the result of a long chain of split-phase I/O is no longer
+required due to a cancellation event, the cancellation callback is itself
+responsible for aborting the ongoing unnecessary operations.
+%
+\CEU relies on the \code{par/or} and \code{finalize} structured mechanisms to
+manage resources automatically.
 
@@
     - which is a lot of time
+
+- artigo synchronous languages
+
+- shim
+
+- electre
 
@@
 
-%% Bibliography
 \bibliography{my,other}
 
-%% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
-
 \end{document}

### 2018-08-04 paper.tex
...

@@
 
+%TODO: energy savings
+%TODO: timers (mux, many impls, energy)
+
 Reactive applications interact continuously and in real time with the external
@@
 %
-The applications built on top of these drivers show significant energy savings
-due to automatic standby.
+The applications built on top of these drivers show optimal energy savings due
+to automatic standby.
 
@@
 and energy-aware runtime of \CEU.
-
-TODO: energy savings
+Figure~\ref{fig.energy} shows our initial tests with five existing C/Arduino
+applications rewritten in \CEU.
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{energy}
+\caption{ Energy consumption in \emph{mA} for five existing C/Arduino
+          applications rewritten in \CEU.
+\label{fig.energy}
+}
+\end{figure}
 
@@
 
+\begin{comment}
 - CRP
@@
 - electre
+\end{comment}
 

### 2018-08-05 paper.tex
WIP

@@
 %
-As illustrated in Figure~\ref{fig.env}, peripherals can be encapsulated as a
+%As illustrated in Figure~\ref{fig.env},
+Peripherals can be encapsulated as a
 single component, \emph{the environment}, which connects with the application
@@
 
+\begin{comment}
 \begin{figure}
@@
 \end{figure}
+\end{comment}
 
@@
 
+\begin{comment}
 Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
@@
 finalization.
+\end{comment}
 
@@
 
+\begin{comment}
 Structured reactive programming lets developers write code in direct style,
@@
 execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
+\end{comment}
 
@@
 %
-The \code{par/or} is regarded as an \emph{orthogonal preemption
-primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
-in order to affect each other.
+%The \code{par/or} is regarded as an \emph{orthogonal preemption
+%primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
+%in order to affect each other.
 
@@
 %
-The event loop in Figure~\ref{fig.env} suggests this behavior since the
+%The event loop in Figure~\ref{fig.env} suggests this behavior since
+The
 environment can only generate a new input after the application yields control
-and closes the loop.
+and closes the event loop.
 %
@@
 
+\begin{comment}
 Figure~\ref{fig.ticks} illustrates the synchronous execution model of \CEU.
@@
 \endgroup
-
+\end{comment}
+
+\begin{comment}
 In order to guarantee responsiveness, the synchronous model relies on the
@@
 responsive to upcoming events.
+\end{comment}
 
@@
 application and the environment even in the presence of asynchronous ISRs.
-Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
-original application represented as the synchronous side and the environment as
-the asynchronous side, all inside the same \CEU application.
+%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
+%original application represented as the synchronous side and the environment as
+%the asynchronous side, all inside the same \CEU application.
 %
@@
 The example illustrates the clear separation between the application and the
-driver through an event-driven interface that reflects the desired architecture
-of Figures~\ref{fig.env}~and~\ref{fig.async}.
+driver through an event-driven interface. % that reflects the desired architecture
+%of Figures~\ref{fig.env}~and~\ref{fig.async}.
 %
@@
 
-\begin{figure}[t!]
+\begin{figure}[t]
 \begin{minipage}[t]{0.50\linewidth}
@@
 
+The addition of asynchronous ISRs poses real threats concerning race conditions
+since they interrupt programs at arbitrary points.
+%
+Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
+that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
+%
+The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
+and a circular buffer to prevent data loss (ln~4).
+The buffer is indispensable because the microcontroller may cope with the USART 
+speed.
+%
+The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
+buffer (ln~9) and signals the application (ln~10).
+%
+As a possible race condition, note that the ISR may fire and append a new byte
+to the buffer (ln~9) just before the application clears it (ln~21), in which
+case the new byte will be lost forever.
+
+\CEU treats an \code{async/isr} as a block of code that runs in parallel with
+the rest of the program (hence the required prefix \code{spawn}).
+%
+This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
+Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
+a compile-time error.
+%
+The programmer is responsible to protect concurrent memory accesses with
+\code{atomic} blocks, which disables interrupts (supposedly) for a short period
+of time.
+
 \begin{lstlisting}[
@@
 \end{lstlisting}
-
-The addition of asynchronous ISRs poses real threats concerning race conditions
-since they interrupt programs at arbitrary points.
-%
-Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
-that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
-%
-The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
-and a circular buffer to prevent data loss (ln~4).
-The buffer is indispensable because the microcontroller may cope with the USART 
-speed.
-%
-The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
-buffer (ln~9) and signals the application (ln~10).
-%
-As a possible race condition, note that the ISR may fire and append a new byte
-to the buffer (ln~9) just before the application clears it (ln~21), in which
-case the new byte will be lost forever.
-
-\CEU treats an \code{async/isr} as a block of code that runs in parallel with
-the rest of the program (hence the required prefix \code{spawn}).
-%
-This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
-Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
-a compile-time error.
-%
-The programmer is responsible to protect concurrent memory accesses with
-\code{atomic} blocks, which disables interrupts (supposedly) for a short period
-of time.
 
@@
 
-Our runtime implements the event loop proposed in Figure~\ref{fig.async}, which
+Our runtime
+%implements the event loop proposed in Figure~\ref{fig.async}, which
 alternates between synchronous, well-behaved execution, and asynchronous,
@@
 and energy-aware runtime of \CEU.
+
+\begin{comment}
 Figure~\ref{fig.energy} shows our initial tests with five existing C/Arduino
@@
 \end{figure}
-
+\end{comment}
+
+\begin{comment}
 \section{Related Work}
@@
 
-\begin{comment}
+%\begin{comment}
 - CRP
@@
 - electre
+%\end{comment}
 \end{comment}
+
+\section{Conclusion}
+
+Drivers
+
+Our initial tests show optimal energy savings for applications in idle states.
 

### 2018-08-06 paper.tex
...

@@
 In this work, we propose an interrupt handler primitive for a reactive language
-targeting embedded systems in order to take control of the whole event loop:
-from a sensor input source and back to an actuator output.
+targeting embedded systems in order to take control of the whole event loop,
+from input to output.
 %
@@
     reactive programming,
-    sleep modes,
     standby,
@@
 single component, \emph{the environment}, which connects with the application
-in an event loop:
+through an event loop:
 the application sits idle until the environment awakes it on the occurrence of
 an input;
-the application reacts to the input and generates back one or more outputs
+the application reacts to the input and generates back outputs
 which affect the environment;
@@
 
-The environment is typically implemented in a host language (e.g., C) and
-controls the main event loop, invoking entry points into the reactive language
-runtime on the occurrence of inputs, and also receiving output calls from it,
-both through a documented API.
+Reactive languages provide dedicated syntax and semantics to deal with events
+and greatly simplify the development of applications in this domain.
+%
+However, the environment is typically implemented in a host language (e.g., C)
+and controls the main event loop, invoking entry points into the reactive
+language runtime on the occurrence of inputs, and also receiving output calls
+from it.
+%, both through a documented API.
 %
@@
 
-The event-based interface between the application and the environment is
+The event-based interface between the reactive language and the environment is
 arguably inevitable and also happens at the appropriate layer, since it
-connects the application with the operating system resources through a
+connects the application with the operating system (OS) resources through a
 non-invasive API.
@@
 In this work, we propose interrupt service routines (ISRs) as an asynchronous
-primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
+primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
 ISRs empower reactive applications to also implement their own device drivers
@@
 microcontrollers, in which applications run in the bare metal, without
-operating system support.
-In this context, device drivers are typically libraries compiled together with
-the main program.
-
-The lack of an OS opens an opportunity to explore a tighter integration between
-the application and its device drivers at the language level, resulting in an
-overall simplicity and tractability of the system.
-%
-In particular, \CEU is already amenable to a simple static analysis to detect
-race conditions which we extended in this work to also encompass ISRs.
+OS support.
+%In this context, device drivers are typically libraries compiled together with
+%the main program.
+%
+%The lack of an OS opens opportunities to explore a tighter integration between
+%the application and its device drivers at the language level, resulting in an
+%overall simplicity and tractability of the system.
+%
+%In particular, \CEU is already amenable to a simple static analysis to detect
+%race conditions which we extended in this work to also encompass ISRs.
 Applications can share data buffers with drivers to avoid unnecessary copying,
@@
 %
-\CEU also provides a lexical finalization mechanism that we adopt in drivers to
+\CEU provides a lexical finalization mechanism that we adopt in drivers to
 properly disable interrupts and turn off peripherals completely.
-For instance, we can delimit the scope of a driver similarly to a
-lexically-scoped variables.
+%For instance, we can delimit the scope of drivers similarly to lexically-scoped
+%variables.
 %
@@
 sleep automatically at optimal sleeping modes after each input reaction.
-
-We implemented an extensible runtime that exposes hooks for the interrupts and
-power manager, which we validated in two microcontrollers: an
-\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
-%
-We also implemented drivers for a variety of peripherals, such as GPIO, A/D
+%
+%We implemented an extensible runtime that exposes hooks for the ISRs and
+%power manager, which we validated in two microcontrollers: an
+%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
+%
+We implemented drivers for a variety of peripherals, such as GPIO, A/D
 converter, USART, SPI, and nRF24L01 transceiver.
@@
 
-\begin{comment}
 Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
-wireless devices, which also provides asynchronous events with a compile-time
-data race detector~\cite{wsn.nesc}, and automatic energy management triggered
-from idle CPU states~\cite{wsn.icem}.
+wireless devices, which also provides asynchronous events with compile-time
+data race detection~\cite{wsn.nesc}, and automatic energy management
+in idle CPU states~\cite{wsn.icem}.
 %
 We adapted these ideas to the structured reactive programming model of \CEU,
-and refined them with stronger guarantees due to lexical scope and automatic
-finalization.
-\end{comment}
+and refined them with stronger guarantees due to its support for lexically
+scoped lines of execution with automatic finalization.
 
@@
 
-\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{ceu.tecs17} reactive
+\CEU~\cite{ceu.sensys13} is a Esterel-based
+%~\cite{ceu.tecs17}
+reactive
 programming language targeting resource-constrained embedded systems, with the
@@
 
-\begin{comment}
 Structured reactive programming lets developers write code in direct style,
 recovering from the inversion of control imposed by event-driven
-execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
-\end{comment}
+execution~\cite{rp.deprecating}. %,rp.rescala,sync_async.cooperative
 
@@
 %
-%The \code{par/or} is regarded as an \emph{orthogonal preemption
-%primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
-%in order to affect each other.
+The \code{par/or} is regarded as an \emph{orthogonal preemption
+primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
+in order to affect each other.
 
@@
 
-Drivers
-
+We extend the synchronous language \CEU with asynchronous interrupt service
+routines (ISRs) to take control of the whole event loop in reactive systems,
+from input to output.
+%
+ISRs empower reactive applications to also implement their own device drivers
+and self-generate inputs, bypassing the need for a host environment.
+%
+However, asynchronous execution confronts the well-behaved semantics of
+synchronous languages with possible race conditions.
+%
+In order to mitigate this threat, we extend the static concurrency checks of
+\CEU to detect data races between ISRs and the application.
+
+Our approach for writing drivers take advantage of the lexical structured
+mechanisms of \CEU, such as parallel compositions and abortion of lines of
+execution, to offer stronger guarantees.
+%
+For instance, the visibility of a driver can be delimited to a scoped block
+with the guarantee that its resources will not leak after usage.
+%
+The same mechanism permits that drivers signal the power manager that its
+peripheral is no longer required, which allows applications to use optimal
+sleep modes to save energy automatically.
+%
 Our initial tests show optimal energy savings for applications in idle states.

### 2018-08-07 paper.tex
...

@@
 targeting embedded systems in order to take control of the whole event loop,
-from input to output.
+from inputs to outputs.
 %
@@
 %As illustrated in Figure~\ref{fig.env},
-Peripherals can be encapsulated as a
+Peripherals can be abstracted in a
 single component, \emph{the environment}, which connects with the application
 through an event loop:
-the application sits idle until the environment awakes it on the occurrence of
+the application sits idle until the environment awakes it on
 an input;
-the application reacts to the input and generates back outputs
+the application reacts and generates back outputs
 which affect the environment;
@@
 Reactive languages provide dedicated syntax and semantics to deal with events
-and greatly simplify the development of applications in this domain.
-%
-However, the environment is typically implemented in a host language (e.g., C)
+and simplify the development of applications in this domain.
+%
+However, the environment is still typically implemented in a host language (e.g., C)
 and controls the main event loop, invoking entry points into the reactive
@@
 %
+\begin{comment}
 As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
@@
 subscriptions~\cite{frp.elm.ports}.
+\end{comment}
 
 The event-based interface between the reactive language and the environment is
-arguably inevitable and also happens at the appropriate layer, since it
-connects the application with the operating system (OS) resources through a
-non-invasive API.
-%
-However, this separation reveals the environment as a rigid system component
+arguably inevitable,
+% and also happens at the appropriate layer, since it
+%connects the application with the operating system (OS) resources through a
+%non-invasive API.
+%
+%However, this separation
+but reveals the environment as a rigid system component
 that evolves in separate from the application.
@@
 Furthermore, in the context of embedded systems, a proper host OS may even be
-absent or lacking enough device drivers, which requires more low-level
+absent,
+% or lacking enough device drivers,
+which requires more low-level
 intervention from the application.
@@
 primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
-ISRs empower reactive applications to also implement their own device drivers
-and self-generate inputs, bypassing the need for a host environment.
-%
-\CEU targets resource-constrained architectures, such as Arduino-compatible
-microcontrollers, in which applications run in the bare metal, without
-OS support.
+ISRs empower reactive applications with their own device drivers which
+self-generate inputs, bypassing the need for a host environment.
+%
+\CEU targets OS-less embedded architectures, such as Arduino-compatible
+microcontrollers. %, in which applications run in the bare metal, without
+%OS support.
 %In this context, device drivers are typically libraries compiled together with
@@
 %race conditions which we extended in this work to also encompass ISRs.
-Applications can share data buffers with drivers to avoid unnecessary copying,
+Applications in \CEU can share data buffers with drivers, to avoid unnecessary copying,
 with some static guarantee that no data races will occur.
 %
-\CEU provides a lexical finalization mechanism that we adopt in drivers to
+\CEU also provides a lexical finalization mechanism % that we adopt in drivers
+to
 properly disable interrupts and turn off peripherals completely.
@@
 to inputs in bounded time and remain in idle states susceptible to standby.
-With the help of drivers and a power manager, we can put the microcontroller to
-sleep automatically at optimal sleeping modes after each input reaction.
+With the help of drivers and a power manager, the microcontroller
+sleeps automatically at optimal sleeping modes after each input reaction.
 %
@@
 We implemented drivers for a variety of peripherals, such as GPIO, A/D
-converter, USART, SPI, and nRF24L01 transceiver.
-%
-The applications built on top of these drivers show optimal energy savings due
-to automatic standby.
+converter, USART, and nRF24L01 transceiver. %, SPI
+%
+%The applications built on top of these drivers show optimal energy savings due
+%to automatic standby.
 
 Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
-wireless devices, which also provides asynchronous events with compile-time
+sensor networks, which also provides asynchronous events with %compile-time
 data race detection~\cite{wsn.nesc}, and automatic energy management
-in idle CPU states~\cite{wsn.icem}.
-%
-We adapted these ideas to the structured reactive programming model of \CEU,
+~\cite{wsn.icem}.
+%
+We adapted these ideas to the structured reactive model of \CEU,
 and refined them with stronger guarantees due to its support for lexically
@@
 
-\CEU~\cite{ceu.sensys13} is a Esterel-based
+\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{esterel.ieee91}
 %~\cite{ceu.tecs17}
@@
 routines (ISRs) to take control of the whole event loop in reactive systems,
-from input to output.
+from inputs to outputs.
 %
@@
 %
-In order to mitigate this threat, we extend the static concurrency checks of
-\CEU to detect data races between ISRs and the application.
-
-Our approach for writing drivers take advantage of the lexical structured
-mechanisms of \CEU, such as parallel compositions and abortion of lines of
-execution, to offer stronger guarantees.
-%
-For instance, the visibility of a driver can be delimited to a scoped block
-with the guarantee that its resources will not leak after usage.
-%
-The same mechanism permits that drivers signal the power manager that its
-peripheral is no longer required, which allows applications to use optimal
+To mitigate this threat, we extend the static concurrency checks of \CEU to
+detect data races between ISRs and the application.
+
+Our approach for writing drivers relies on the lexical structured mechanisms of
+\CEU, such as parallel compositions and abortion of lines of execution, to
+offer stronger guarantees.
+%
+For instance, the visibility of drivers can be delimited to scoped blocks to
+avoid resource leaks after usage.
+%
+The same mechanism allows drivers to signal the power manager that associated
+peripherals are no longer required, allowing applications to use optimal
 sleep modes to save energy automatically.

### 2018-08-07 paper.tex
...

@@
   float=h,
-  caption={A \CEU program that toggles the state of the LED whenever a radio
+  caption={A \CEU program that blinks the LED whenever a radio
            packet is received, terminating on a button press, always with the
@@
 The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
-trail terminates (e.g., from a button press), the LED will be unconditionally
+trail terminates (e.g., from a button press), the LED is unconditionally
 turned off~(ln~8).
@@
 the \code{await} and \code{emit} primitives, which awaits an input event and
-generates an output event, respectively.
+generates an output event, respectively (ln~5,8,11--14 in the example).
 
@@
 will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
-off unconditionally (ln~8), since its enclosing block goes out of scope.
+off (ln~8) since its enclosing block goes out of scope.
 %
@@
 primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
-in order to affect each other.
+to affect each other.
 
@@
 
-\subsection{Lexically-Scoped Interrupt Service Routines}
+\subsection{Scoped Interrupt Service Routines}
 \label{sec.isrs}
@@
 %
-An ISR starts to execute as soon as the hardware interrupt occurs, suspending
+An ISR starts to execute as soon as a hardware interrupt occurs, suspending
 the ongoing program flow abruptly.
 Such asynchronous behavior reflects the inherent concurrent nature of
-peripherals interacting with the external world.
+peripherals interacting with the real world.
 
@@
 the assumption that reactions are atomic no longer holds.
-Not only this may lead to race conditions at a fine grain, but may also affect
-the ordering of events at a coarse grain: the effect of an earlier event may be
-perceived after the effect of a later event.
+Not only this might lead to race conditions at a fine grain, but might also
+affect the ordering of events at a coarse grain: the effect of an earlier event
+might be perceived after the effect of a later event.
 %
@@
 ]
-output high/low PIN_13;  // connected to an LED
-input  high/low PIN_02;  // connected to a button
+output high/low PIN_13;  // connected to LED
+input  high/low PIN_02;  // connected to button
 #include "gpio.ceu"      // GPIO driver
@@
 This allows drivers to take advantage of existing libraries of embedded
-toolchains, such as Arduino.
+toolchains such as Arduino.
 %
@@
 %
-An ISR in \CEU will typically emit an input event inside its body to awake the
-synchronous side (ln~17).
+An ISR in \CEU will typically perform simple operations and emit an input event
+inside its body to awake the synchronous side (ln~17).
 However, although the ISR executes asynchronously when the interrupt occurs,
@@
   float=t,
-  caption={The A/D driver (ln~1--14) is included in the application inside a
-           lexically-scoped block (ln~20--24), which turns off all driver
-           functionalities automatically on termination.},
+  caption={The A/D driver (ln~1--14) is included in the application (ln~21)
+           inside a lexically-scoped block (ln~20--24), which turns off all
+           driver functionalities automatically on termination.
+           \emph{(The driver and application are actually in separate files.)}},
   label={lst.adc},
@@
 
-The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
-driver for the A/D converter.
-Note that the driver (ln~1--14) and application (ln~16--27) are actually in
-separate files.
+The next example in Listing~\ref{lst.adc} illustrates the use of a
+lexically-scoped driver for the A/D converter.
 %
@@
 unconditionally whenever its enclosing block goes out of scope (ln~24).
-The finalize completely disables all A/D functionality to save energy.
+The finalize completely disables all A/D functionality to save energy
+(ln~8--10).
 %
 As discussed in the previous section, the finalizer would also execute if
-aborted from a \code{par/or} enclosing the driver.
-%
-Note that the driver interface (ln~3--4) is visible only inside the block and
-thus cannot be inadvertently used outside it.
+aborted from a hypothetical \code{par/or} enclosing the driver.
+%
+Note that the driver interface (ln~3--4) is visible only inside the block
+(ln~20--24) and thus cannot be inadvertently used outside it.
 
@@
 
-The addition of asynchronous ISRs poses real threats concerning race conditions
+The addition of asynchronous ISRs now poses real threats concerning race conditions
 since they interrupt programs at arbitrary points.
@@
 The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
-and a circular buffer to prevent data loss (ln~4).
-The buffer is indispensable because the microcontroller may cope with the USART 
+and a buffer to prevent data loss (ln~4).
+The buffer is indispensable because the microcontroller might not cope with the USART 
 speed.
@@
 
-\CEU treats an \code{async/isr} as a block of code that runs in parallel with
-the rest of the program (hence the required prefix \code{spawn}).
+\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
+the rest of the program (hence the prefix \code{spawn}).
 %
@@
 The programmer is responsible to protect concurrent memory accesses with
-\code{atomic} blocks, which disables interrupts (supposedly) for a short period
-of time.
+\code{atomic} blocks, which disables interrupts for a short period
+of time (supposedly).
 
@@
            resulting in a pontential race condition.
-           \emph{(The driver and application code are in separate files.)}},
+           \emph{(The driver and application are actually in separate files.)}},
   label={lst.usart},
@@
 input none USART_RX;
-var[32*] byte rx_buf;             // `*' = circular
+var[32] byte rx_buf;
 
@@
            to the application.
-           \emph{(The driver and application code are in separate files.)}},
+           \emph{(The driver and application are actually in separate files.)}},
   label={lst.usart.code},
@@
 input none USART_RX;
-var[32*] byte rx_buf;
+var[32] byte rx_buf;
 
@@
 
-Our runtime
+The language runtime
 %implements the event loop proposed in Figure~\ref{fig.async}, which
@@
 Since the language now has full control over the event loop and only executes
-from interrupts, it is possible for the runtime to enter in sleep mode
+from interrupt requests, it is possible for the runtime to enter in sleep mode
 automatically to save energy.
-
+%
 Listing~\ref{lst.env} is a realization of this architecture:
@@
 which starts the program, spawns the ISRs, and reaches the first \code{await}
-statements in the multiple lines of code.
-%
-Then, the runtime enters an infinite loop (ln~4--11) that queries the input
+statements in the multiple lines of code of the program.
+%
+Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
 queue (ln~6) to awake the program (ln~7).
 %
-Note that ISRs execute asynchronously with the loop, but they typically only
+Note that ISRs execute asynchronously with the loop and
 emit an input event into the queue which will be queried in a subsequent
@@
 %
-As an example, the least efficient mode of the \emph{ATmega328P} allows for
+As an example, the least efficient mode of the
+\emph{ATmega328P}~\cite{arduino.atmega328p} allows for
 timer interrupts since it keeps its internal clock active, while the most
 efficient mode turns off all peripherals and can only awake from external
-interrupts.
+interrupts (e.g., GPIO falling edge).
 
@@
 %
-The third configuration keeps a bit vector of the active drivers during runtime
+The third sleep configuration keeps a bit vector of the active drivers during runtime
 and chooses the most efficient mode, querying this vector every time
@@
   float=t,
-  caption={Interaction between the power manager, USART driver, and application
-           (each code is actually in a separate file).},
+  caption={Interaction between the power manager, USART driver, and application.
+           \emph{(Each code is actually in a separate file.)}},
   label={lst.pm},
@@
 %
-The application is a loop that awaits in parallel for a button press (ln~36) or
+The application is a loop that initially awaits in parallel for a button press (ln~36) or
 receiving 10 bytes (ln~39).
-Let's call this initial state \code{STATE-A}.
+Let's call this state \code{STATE-A}.
 The \code{par/or} terminates when either of them occurs, going to the next line
@@
 Let's call this other state \code{STATE-B}.
-After the button click, the program loops back to \code{STATE-A}.
+After another button click, the program loops back to \code{STATE-A}.
 %
@@
 The USART driver needs to be extended with calls to enable and disable the
-USART state inside the power manager: just before awaiting, the driver enables
+USART state inside the power manager: just before awaiting (ln~28), the driver enables
 the USART (ln~24) and creates a finalization block to unregister it on
@@
 
-Note that applications need not to be explicit about energy management to take
+Note that applications never need to be explicit about energy management to take
 advantage of sleep modes.
-All happens automatically because of the synchronous-reactive execution model
-and energy-aware runtime of \CEU.
+All happens automatically because of the synchronous reactive execution model
+and the interaction between the drivers and energy-aware runtime of \CEU.
 
@@
 %
-To mitigate this threat, we extend the static concurrency checks of \CEU to
+To mitigate this threat, we extend the static concurrency checks of \CEU to also
 detect data races between ISRs and the application.
 
-Our approach for writing drivers relies on the lexical structured mechanisms of
+Our approach for developing drivers relies on the lexical structured mechanisms of
 \CEU, such as parallel compositions and abortion of lines of execution, to
@@
 For instance, the visibility of drivers can be delimited to scoped blocks to
-avoid resource leaks after usage.
+avoid resource leaks.
 %
@@
 peripherals are no longer required, allowing applications to use optimal
-sleep modes to save energy automatically.
+sleep modes automatically.
 %

### 2018-08-09 paper.tex
...

@@
   morekeywords={%
-    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
-    await, bool, call, class, data, define, deterministic, do, each, else,
-    emit, end, escape, event, every, finalize, hor, implementation, in,
-    input, interface, loop, min, native, new, nohold, not, or, output, par,
-    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
-    code, public, private,
-    var, watching, when, with},
+    NEVER, FOREVER, and, high, low, async, isr, include, byte, none, atomic,
+    await, bool, call, data, define, deterministic, do, else,
+    emit, end, escape, event, every, finalize, in,
+    input, loop, native, new, nohold, not, or, output, par,
+    pool, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
+    code, var, watching, with},
 }
@@
 targeting embedded systems in order to take control of the whole event loop,
-from inputs to outputs.
+from input generation up to output effects.
 %
@@
 %
-It also requires two languages and the programmer has to deal with multiple
+It also requires the programmer to deal with multiple
 syntaxes, incompatible type systems, and different address spaces.
 %
-Furthermore, in the context of embedded systems, a proper host OS may even be
-absent,
-% or lacking enough device drivers,
+Furthermore, in the context of embedded systems, a proper host operating system (OS) may even be
+absent or lacking enough device drivers,
 which requires more low-level
@@
 \CEU also provides a lexical finalization mechanism % that we adopt in drivers
-to
-properly disable interrupts and turn off peripherals completely.
+which
+safely disables drivers and turns off peripherals completely to save energy.
 %For instance, we can delimit the scope of drivers similarly to lexically-scoped
@@
 We implemented drivers for a variety of peripherals, such as GPIO, A/D
-converter, USART, and nRF24L01 transceiver. %, SPI
+converter, USART, SPI, and an RF transceiver.
 %
@@
 sensor networks, which also provides asynchronous events with %compile-time
-data race detection~\cite{wsn.nesc}, and automatic energy management
-~\cite{wsn.icem}.
+data race detection~\cite{wsn.nesc} and automatic energy
+management~\cite{wsn.icem}.
 %
@@
 recovering from the inversion of control imposed by event-driven
-execution~\cite{rp.deprecating}. %,rp.rescala,sync_async.cooperative
+execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
 
@@
 The synchronous execution model of \CEU dictates that reactions to input events
-are atomic and that incoming events are never lost.
-%
-%The event loop in Figure~\ref{fig.env} suggests this behavior since
-The
-environment can only generate a new input after the application yields control
-and closes the event loop.
+are atomic and that incoming events are never lost, which we refer to as the
+\emph{atomicity} and \emph{responsiveness} properties, respectively.
+%
+%The event loop in Figure~\ref{fig.env} suggests this behavior since the
+%environment can only generate a new input after the application yields control
+%and closes the event loop.
 %
 In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
-environment, the synchronous model ensures atomicity and responsiveness in the
-application:
+environment, the synchronous model ensures these properties by adopting an
+event queue:
     the first \code{await} awakes and turns the LED off atomically
     (ln~12--13);
-    and the second \code{await} will awake in sequence (ln~14).
+    only after that, the second \code{await} will awake (ln~14).
 %
@@
 %
-Our idea is to push all subtleties of asynchronous execution into device
-drivers, which emit input events to regular synchronous code in \CEU, exactly
-as before.
+Our approach is to push all subtleties of asynchronous execution into device
+drivers, which impersonate the environment and are responsible to emit input
+events towards regular synchronous code.
+%
+Synchronous code preserves the atomicity and responsiveness properties,
+communicating with the environment only through events, exactly as before.
+
+As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
+GPIOs to connect a button to an LED via software such that the LED (pin 13) is
+on whenever the button (pin 02) is pressed, and off whenever it is unpressed.
+
+Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
+with the external world and includes the asynchronous side as a driver
+(ln~1--3).
+It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
+whenever the button changes (ln~7), toggles the state of the LED with the new
+value (ln~8).
 
@@
 
-As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
-GPIOs to connect a button (pin 02) to an LED (pin 13) via software such that
-the LED is on whenever the button is pressed and off whenever it is unpressed.
-
-Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
-with the external world and includes the asynchronous side as a driver
-(ln~1--3).
-It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
-whenever the button changes (ln~7), toggles the state of the LED with the new
-value (ln~8).
-
 Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
@@
 An ISR in \CEU will typically perform simple operations and emit an input event
-inside its body to awake the synchronous side (ln~17).
+to awake the synchronous side (ln~17).
 However, although the ISR executes asynchronously when the interrupt occurs,
-the \code{emit} goes into a queue (to be discussed) and does not interrupt the
-ongoing (suspended) reaction on the synchronous side.
-%
-In the example, the input driver also configures pin 2 to behave as input and
-to generate external interrupts on level transitions (ln~11--15).
+the \code{emit} goes into a queue and does not affect the synchronous side
+immediately.
+%
+In the example, when the input driver is included, it also configures pin 2 to
+behave as input and to generate external interrupts on level transitions
+(ln~11--15).
 
@@
 
-  output int ADC_REQUEST;
-  input  int ADC_DONE;
+  output int ADC_REQUEST;   // requests a conversion
+  input  int ADC_DONE;      // conversion is done
 
@@
 unconditionally whenever its enclosing block goes out of scope (ln~24).
-The finalize completely disables all A/D functionality to save energy
+The \code{finalize} completely disables all A/D functionality to save energy
 (ln~8--10).
@@
 
+\newpage
 \subsection{Safe Shared-Memory Concurrency}
@@
 
-In \CEU, when multiple trails awake in the same reaction, they execute in
-lexical order, i.e., in the order they appear in the source code.
-%
-In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
+In \CEU, when multiple trails awake in the same reaction, they execute
+atomically in lexical order, i.e., in the order they appear in the source code.
+%
+In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) and
 assign to it in parallel trails (ln~6, 9).
@@
 reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
-occur simultaneously in accordance with the synchronous model.
+occur simultaneously due to the atomicity property of the synchronous model.
 %Hence, for the sequence of events \code{A->B}, \code{x} becomes
@@
 because they execute in reaction to the same event \code{A}.
-Nevertheless, because \CEU follows lexical order, the execution is still
-deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
+Nevertheless, because \CEU follows lexical order and executes atomically, the
+outcome is still deterministic, and \code{y} always becomes \code{4}
+(\code{(1+1)*2}).
 %
@@
 This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
-Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
-a compile-time error.
+Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
+raises a compile-time error.
 %
@@
 \code{atomic} blocks, which disables interrupts for a short period
-of time (supposedly).
+of time.
 
@@
 
-input none USART_RX;
-var[32] byte rx_buf;
+input none USART_RX;    // data is available
+var[32] byte rx_buf;    // data buffer
 
@@
 \CEU supports reactive abstractions (similar to coroutines) that help hiding
-the concurrency complexity inside the driver.
+concurrency issues inside drivers.
 %
@@
 (ln~7--8) and implements the code abstraction (ln~15--27) that protects the
-access to the shared buffer with an \code{atomic} block (ln~18--21).
+accesses to the shared buffer with an \code{atomic} block (ln~18--21).
 %
@@
 
+\newpage
 \subsection{Energy-Efficient Runtime}
@@
     evt_t evt;
-    if (ceu_input(&evt)) {  // query input queue
-      ceu_sync(&evt);       // execute synchronous
+    if (ceu_input(&evt)) {  // queries input queue
+      ceu_sync(&evt);       // executes synchronous
     } else {
@@
 which starts the program, spawns the ISRs, and reaches the first \code{await}
-statements in the multiple lines of code of the program.
+statements in the multiple trails of the program.
 %
@@
 %
-Note that ISRs execute asynchronously with the loop and
-emit an input event into the queue which will be queried in a subsequent
-iteration.
-%
-Note also that if the queue is empty, the runtime enters in sleep mode to save
+ISRs execute asynchronously, emitting input events into the queue, which will
+be queried in subsequent iterations of the loop.
+%
+Note that if the queue is empty, the runtime enters in sleep mode to save
 energy (ln~9), and will only awake on a new hardware interrupt.
@@
 The third configuration achieves optimal efficiency but requires a tight
-interaction between the drivers and the power management runtime.
+interaction between the drivers and the power manager.
 %
@@
 %
-The USART driver needs to be extended with calls to enable and disable the
+The USART driver now needs to be extended with calls to enable and disable the
 USART state inside the power manager: just before awaiting (ln~28), the driver enables
-the USART (ln~24) and creates a finalization block to unregister it on
+the USART (ln~24) and creates a finalization block to disable it on
 termination (ln~25--27).
@@
 routines (ISRs) to take control of the whole event loop in reactive systems,
-from inputs to outputs.
+from input generation up to output effects.
 %

### 2018-08-12 paper.tex
(>) Sugestoes Sztajnberg

@@
 In reactive and event-based systems, execution is guided by an external
-environment that generates inputs to the application and consumes outputs from
-it.
+environment that generates inputs to the application and is affected by outputs
+from it.
 Reactive languages provide dedicated syntax and semantics to deal with events
@@
 In this work, we propose an interrupt handler primitive for a reactive language
-targeting embedded systems in order to take control of the whole event loop,
+that targets embedded systems in order to take control of the whole event loop,
 from input generation up to output effects.
@@
 We propose the new asynchronous primitive in the context of the synchronous
-language \CEU and discuss how they synergize to avoid race conditions at
-compile time, support lexically-scoped drivers, and provide automatic standby
-for applications.
+language \CEU and discuss how they synergize to prevent runtime race conditions
+at compile time, support lexically-scoped drivers, and provide automatic
+standby for applications.
 %and hot swapping of device drivers
@@
 an input;
-the application reacts and generates back outputs
-which affect the environment;
+the application reacts and generates outputs back, which affect the environment;
 the application becomes idle and the loop restarts.
@@
 
-Reactive languages provide dedicated syntax and semantics to deal with events
+Reactive languages provide syntax and semantics specialized to deal with events
 and simplify the development of applications in this domain.
@@
 primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
-ISRs empower reactive applications with their own device drivers which
+ISRs empower reactive applications with their own device drivers that
 self-generate inputs, bypassing the need for a host environment.
@@
 \CEU also provides a lexical finalization mechanism % that we adopt in drivers
-which
-safely disables drivers and turns off peripherals completely to save energy.
+that safely disables drivers and turns off peripherals to save energy.
 %For instance, we can delimit the scope of drivers similarly to lexically-scoped
@@
   float=t,
-  caption={Asynchronous driver for GPIO which, on hardware interrupts, emits an
+  caption={Asynchronous driver for GPIO that, on hardware interrupts, emits an
            input event into a queue.},
@@
 %
-The driver specifies a finalization code (ln~6--12) which executes
+The driver specifies a finalization code (ln~6--12) that executes
 unconditionally whenever its enclosing block goes out of scope (ln~24).
@@
 %
-The first configuration is a \emph{nop} which simply keeps the event loop
+The first configuration is a \emph{nop} that simply keeps the event loop
 running all the time without ever sleeping.

### 2018-08-15 paper.tex
(>) Revisões Sztajnberg até 2.1

@@
 
-\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{esterel.ieee91}
+\CEU~\cite{ceu.sensys13} is an Esterel-based~\cite{esterel.ieee91}
 %~\cite{ceu.tecs17}
@@
 \begin{description}
-\item [Reactive:] code only executes in reactions to events and is idle most of
+\item [Reactive:] code only executes in reaction to events and is idle most of
     the time.
@@
 Structured reactive programming lets developers write code in direct style,
-recovering from the inversion of control imposed by event-driven
+reversing from the inversion of control imposed by event-driven
 execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
@@
 The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
-trail terminates (e.g., from a button press), the LED is unconditionally
-turned off~(ln~8).
+trail terminates or aborts (i.e., even from a button press), the LED is
+unconditionally turned off~(ln~8).
 
@@
 
+% AS: Agora entendi ... mas não é muito "interweaved" ou seja ... um trail
+% afeta o outro, e voce tira vantagem disso para que a clasula finalize seja
+% afetada ... não seria o caso de deixar claro que a clausula finalize está
+% fora de qalquer trail? Ou por outra, seu eu tivesse vários trails, todos
+% seriam interrompidos. E se cada um tivesse um finalize? todos seriam
+% executados? em que ordem?
+% FS: A questão é justamente ser menos "interweaved": Se eu escrevi o
+% comportamento do LED em uma trilha, essa mesma trilha tem que ser responsável
+% por terminar em um estado seguro para o LED (ou qq recurso). Assim, você não
+% precisa se preocupar em como as outras trilhas executam ou terminam, o seu
+% estado final só depende de você.
+% Se eu colocar para fora de qualquer trilha, eu vou ter que misturar o código
+% com diversos finalizes que não estão relacionados. Ainda tem o problema de
+% saber *onde* colocar. Esse exemplo só tem um par/or, então é óbivo, mas tem
+% programas com loops, par/or, par/and, etc aninhados.
+% Se cada trilha tiver um finalize e todas forem abortadas, elas executam em
+% ordem léxica, que é a mesma regra para quando trilhas acordam na mesma
+% reação.
+
 The synchronous execution model of \CEU dictates that reactions to input events
@@
 event queue:
-    the first \code{await} awakes and turns the LED off atomically
-    (ln~12--13);
-    only after that, the second \code{await} will awake (ln~14).
+    the first \code{await} awakes from the first packet and turns the LED off
+    atomically (ln~12--13);
+    only after that, the second \code{await} will awake from the second packet
+    (ln~14).
+From the point of view of the program, the packets arrive in sequence and each
+is handled in separate.
 %
@@
 
-As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
+As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} use
 GPIOs to connect a button to an LED via software such that the LED (pin 13) is
 on whenever the button (pin 02) is pressed, and off whenever it is unpressed.
+%
+The idea is to use Listing~\ref{lst.gpio.sync} handle synchronous/atomic
+changes in the button to affect the LED, and Listing~\ref{lst.gpio.async} to
+handle asynchronous changes that are enqueued towards the synchronous side.
 
@@
 value (ln~8).
+%
+This code should be concise and straightforward: each button change toggles the
+LED (atomicity) and no button changes are ever lost (responsiveness).
 
@@
 Nevertheless, drivers are typically write-once components developed by embedded
-specialists, which are reused in most applications.
+systems specialists, which are reused in most applications.
 %
@@
 
-  do finalize with
+  finalize with
     {
@@
 
-\newpage
 \subsection{Safe Shared-Memory Concurrency}
@@
 
-\newpage
 \subsection{Energy-Efficient Runtime}

### 2018-08-15 paper.tex
(>) Revisões Sztajnberg 2.2 e Conclusão

@@
 %
-In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) and
-assign to it in parallel trails (ln~6, 9).
+In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) which
+is assigned in parallel trails (ln~6, 9).
 %
@@
 
-The addition of asynchronous ISRs now poses real threats concerning race conditions
-since they interrupt programs at arbitrary points.
+The addition of asynchronous ISRs now poses real threats concerning race
+conditions since they interrupt programs at arbitrary points and could possibly
+share variables with synchronous code.
 %
@@
 spawn async/isr [USART_RX_vect] do
-    rx_buf = rx_buf .. [{UDR0}];
-    emit USART_RX;
+    rx_buf = rx_buf .. [{UDR0}];  // appends new byte
+    if $rx_buf == 1 then
+        emit USART_RX; // emits unless already pending
+    end                // (assumes client consumes all)
 end
@@
     var int i;
-    loop i in [0 -> $rx_buf[ do   // `$' = length ($)
+    loop i in [0 -> $rx_buf[ do   // `$' = length
         // uses rx_buf[i]
@@
 and the interaction between the drivers and energy-aware runtime of \CEU.
+%
+Our initial tests show optimal energy savings for applications in idle
+states~\cite{ceu.lctes18.short}.
 
@@
 sleep modes automatically.
-%
-Our initial tests show optimal energy savings for applications in idle states.
 

### 2018-08-15 paper.tex
(>) Revisões Sztajnberg 2.3

@@
 %
-Since the language now has full control over the event loop and only executes
-from interrupt requests, it is possible for the runtime to enter in sleep mode
-automatically to save energy.
-%
-Listing~\ref{lst.env} is a realization of this architecture:
+The runtime now also has full control over the event loop and only executes
+from interrupt requests.
+This allows us to propose a runtime that enters in sleep mode automatically to
+save energy, as depicted in Listing~\ref{lst.env}:
 %
@@
 Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
-(ln~9):
+(ln~9).
 %
@@
     par/or do
-        await PIN_02;
+        await PIN_02;               // STATE-A
     with
         var[255] byte buf;
-        await Usart_RX(&buf, 10);
+        await Usart_RX(&buf, 10);   // STATE-A
         // uses buf
     end
-    await PIN_02;
+    await PIN_02;                   // STATE-B
 end

### 2018-08-15 paper.tex
(>) intro

@@
 scoped lines of execution with automatic finalization.
+
+\begin{comment}
+We start with an introduction of the structured synchronous reactive
+programming model of \CEU (Section~\ref{sec.ceu}).
+Then, we propose to extend \CEU with asynchronous ISRs and discuss how it can
+interoperate with synchronous code through an event-driven interface
+(Section~\ref{sec.isrs}).
+We also propose a simple static analysis to prevent race conditions between
+synchronous and asynchronous code at compile time (Section~\ref{sec.shared}).
+Finally, we propose a new language runtime that enters in sleep mode
+automatically to save energy (Section~\ref{sec.energy}).
+\end{comment}
 
@@
 \subsection{Safe Shared-Memory Concurrency}
-\label{sec.ceu.shared}
+\label{sec.shared}
 

### 2018-08-16 paper.tex
(>) shrink text

@@
 
+%\usepackage{microtype}
 \usepackage{xspace}
@@
 
+\setlength{\belowcaptionskip}{-1pt}
+
 \begin{document}
@@
 
-\begin{comment}
 We start with an introduction of the structured synchronous reactive
 programming model of \CEU (Section~\ref{sec.ceu}).
-Then, we propose to extend \CEU with asynchronous ISRs and discuss how it can
+Then, we propose to extend it with asynchronous ISRs and discuss how to
 interoperate with synchronous code through an event-driven interface
@@
 automatically to save energy (Section~\ref{sec.energy}).
-\end{comment}
 
@@
 %
-Still, our goal is to preserve the well-behaved interaction between the \CEU
-application and the environment even in the presence of asynchronous ISRs.
+Still, our goal is to preserve the well-behaved interaction between \CEU and
+the environment even in the presence of asynchronous ISRs.
 %Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
@@
   float=t,
-  caption={Synchronous application that turns the LED on whenever the button
-           is pressed and off whenever it is unpressed.},
+  caption={Synchronous code that turns the LED on whenever the button is
+           pressed and off whenever it is unpressed.},
   label={lst.gpio.sync},
@@
 The addition of asynchronous ISRs now poses real threats concerning race
-conditions since they interrupt programs at arbitrary points and could possibly
-share variables with synchronous code.
+conditions since they interrupt execution at arbitrary points and could
+possibly share memory with synchronous code.
 %
@@
 and a buffer to prevent data loss (ln~4).
-The buffer is indispensable because the microcontroller might not cope with the USART 
-speed.
+The buffer is indispensable because the microcontroller might not cope with the
+USART speed.
 %
@@
 %
-As a possible race condition, note that the ISR may fire and append a new byte
-to the buffer (ln~9) just before the application clears it (ln~21), in which
-case the new byte will be lost forever.
-
-\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
-the rest of the program (hence the prefix \code{spawn}).
+As a possible race condition, the ISR might fire and append a new byte to the
+buffer (ln~9) just before the application clears it (ln~21), in which case the
+new byte would be lost forever.
+
+\CEU treats a \code{spawn~async/isr} as a block that runs in parallel with the
+rest of the program (hence the prefix \code{spawn}).
 %
@@
 resolve race conditions.
-\CEU supports reactive abstractions (similar to coroutines) that help hiding
+\CEU provides reactive abstractions (similar to coroutines) that help hiding
 concurrency issues inside drivers.
@@
 receive.
-Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
-local buffer (ln~32).
-%
-The driver (ln~5--27) now hides the low level interface of the previous example
-(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
-accesses to the shared buffer with an \code{atomic} block (ln~18--21).
-%
-The abstraction copies the driver buffer into the application buffer (ln~19)
-up to the requested number of bytes (ln~22).
+Now the application (ln~29--35) invokes the abstraction (ln~33) passing a local
+buffer (ln~32).
+%
+The driver (ln~5--27) now hides the low-level interface (ln~7--8) and
+implements the abstraction (ln~15--27) that protects the accesses to the shared
+buffer with an \code{atomic} block (ln~18--21).
+%
+The code copies the driver buffer into the application buffer (ln~19) up to the
+requested number of bytes (ln~22).
 %
@@
 
-The language runtime
+The language runtime now
 %implements the event loop proposed in Figure~\ref{fig.async}, which
@@
 %
-The runtime now also has full control over the event loop and only executes
-from interrupt requests.
-This allows us to propose a runtime that enters in sleep mode automatically to
-save energy, as depicted in Listing~\ref{lst.env}:
-%
-The runtime defines an input queue (ln~1) in which ISRs enqueue new events
-(e.g., Listing~\ref{lst.usart.code}, ln~12).
-%
-In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
-which starts the program, spawns the ISRs, and reaches the first \code{await}
+Since programs only execute from interrupts, the runtime can now also enter in
+sleep mode automatically to save energy, as depicted in Listing~\ref{lst.env}:
+%
+It first defines an input queue (ln~1) in which ISRs enqueue new events (e.g.,
+Listing~\ref{lst.usart.code}, ln~12).
+%
+The \code{main} function generates thes \emph{boot reaction} once to start the
+program (ln~3), which will spawn the ISRs and reach the first \code{await}
 statements in the multiple trails of the program.
 %
-Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
+Then, the runtime enters the event loop (ln~4--11), which queries the input
 queue (ln~6) to awake the program (ln~7).
@@
 
-Microcontrollers typically support multiple levels of sleep mode, each
+Microcontrollers typically support multiple levels of sleep modes, each
 progressively saves more energy at the expense of keeping less functionality

### 2018-09-27 paper.tex
(>) suggestions from reviews

--- 2018-08-15 paper.tex
@@
 
-\subsection{Scoped Interrupt Service Routines}
+\section{Interrupt Service Routines in \CEU}
+\label{sec.contributions}
+
+In Listing~\ref{lst.ceu}, the input \& output events are external to the
+program and are generated by the host language representing the environment.
+%
+We now propose to extend \CEU with interrupt services routines (ISRs) to take
+control of the whole event loop using a safe and energy-efficient approach.
+
+\subsection{Scoped ISRs}
 \label{sec.isrs}
@@
 output (high/low v) PIN_13 do
-    { digitalWrite(13, @v); }
+    { digitalWrite(13, @(v as int)); }
 end
@@
     pinMode(2, INPUT_PULLUP);
-    EICRA |= (1 << ISC00);  // sets INT0
-    EIMSK |= (1 << INT0);   // turns on INT0
+    EIMSK |= (1 << INT0);  // enables pin-02 ints
+    EICRA |= (1 << ISC00); // for level changes
 }
 spawn async/isr [INT0_vect] do
-    emit PIN_02({digitalRead(2)});
+    emit PIN_02({digitalRead(2)} as high/low);
 end
@@
 executes atomically.
-\CEU supports inline C between curly braces (ln~3,5) with interpolation to
+
+\CEU is designed to interoperate seamlessly with C and supports a compatible
+type system with self-evident conversion rules and same internal representation
+(e.g., \code{high/low} is a boolean, \code{u8} is an unsigned 8-bit value,
+etc.).
+\CEU also supports inline C between curly braces (ln~3,5) and interpolation to
 evaluate \CEU expressions (e.g., \code{@v}).
@@
 toolchains such as Arduino.
-%
+Our drivers still use C, but only a convenient subset (e.g., assignments and
+getters \& setters) for low-level port manipulation.
+
 In the example, when the driver is included, it configures pin 13 as output
@@
 However, although the ISR executes asynchronously when the interrupt occurs,
-the \code{emit} goes into a queue and does not affect the synchronous side
-immediately.
-%
-In the example, when the input driver is included, it also configures pin 2 to
-behave as input and to generate external interrupts on level transitions
-(ln~11--15).
+the \code{emit} goes into an input queue and does not affect the synchronous
+side immediately.
+%
+The input driver also configures pin 2 to behave as input and to generate
+external interrupts on level transitions (ln~11--15).
+%
+The queue is static and new emissions of an unhandled input override the same
+position in the queue.
+This semantics is comparable to interrupt flags in microcontrollers and
+prevents buffer overflows.
 
@@
 %
-Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
-that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
+Listing~\ref{lst.usart} illustrates a minimum serial communication (USART)
+application (ln~15--24) that consumes incoming bytes (ln~19--23) as they arrive
+(ln~18).
 %
@@
 %
-The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
-buffer (ln~9) and signals the application (ln~10).
+The (unsafe) driver ISR (ln~6--13) simply appends incoming bytes to the end of the
+buffer (ln~9) and signals the application (ln~11).
 %
 As a possible race condition, note that the ISR may fire and append a new byte
-to the buffer (ln~9) just before the application clears it (ln~21), in which
+to the buffer (ln~9) just before the application clears it (ln~23), in which
 case the new byte will be lost forever.
@@
 This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
-Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
+Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,23), and \CEU
 raises a compile-time error.
@@
 spawn async/isr [USART_RX_vect] do
-    rx_buf = rx_buf .. [{UDR0}];  // appends new byte
+    rx_buf = rx_buf .. [{UDR0} as byte];  // appends
     if $rx_buf == 1 then
@@
 resolve race conditions.
-\CEU supports reactive abstractions (similar to coroutines) that help hiding
+\CEU supports \code{code} reactive abstractions (similar to coroutines) that help hiding
 concurrency issues inside drivers.

### 2018-09-27 paper.tex
(>) format, authors, ack

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 %% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\author{Francisco Sant'Anna}
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
-}
-\email{first1.last1@inst1.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+  \department{Dep. de Inform\'atica e Ci\^encia da Computa\c{c}\~ao}
+  \institution{Rio de Janeiro State University, Brazil}
+  %\country{Brazil}
+}
+\email{francisco@ime.uerj.br}
+
+\author{Alexandre Sztajnberg}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
+  \department{Dep. de Inform\'atica e Ci\^encia da Computa\c{c}\~ao}
+  \institution{Rio de Janeiro State University, Brazil}
+  %\country{Brazil}
+}
+\email{alexszt@ime.uerj.br}
 
@@
   %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+This work was supported by the Serrapilheira Institute (grant number
+Serra-1708-15612).
 \end{acks}

### 2018-09-27 paper.tex

@@
 %
-As a possible race condition, note that the ISR may fire and append a new byte
-to the buffer (ln~9) just before the application clears it (ln~23), in which
-case the new byte will be lost forever.
-
-\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
-the rest of the program (hence the prefix \code{spawn}).
+As a possible race condition, the ISR might fire and append a new byte to the
+buffer (ln~9) just before the application clears it (ln~23), in which case the
+new byte would be lost forever.
+
+\CEU treats a \code{spawn~async/isr} as a block that runs in parallel with the
+rest of the program (hence the prefix \code{spawn}).
 %
@@
 resolve race conditions.
-\CEU supports \code{code} reactive abstractions (similar to coroutines) that help hiding
+\CEU provides \code{code} reactive abstractions (similar to coroutines) that help hiding
 concurrency issues inside drivers.
@@
   %% extraction tools.
-This work was supported by the Serrapilheira Institute (grant number
-Serra-1708-15612).
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
 \end{acks}
